### 2025-01-31 products/manuscript/Manuscript.qmd
project part 1

---
title: "Exploring clinical and non-clinical Listeria monocytogenes isolates "
author: "Alexis Gonzalez"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

-   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
-   Second Author$^{3, *}$
-   Third Autor$^{3, \dagger}$
-   Last Author$^{1, 2, \land}$

**Author affiliations**

1.  College of Public Health, University of Georgia, Athens, GA, USA.
2.  Another Affiliation, Athens, GA, USA.
3.  Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: some\@email.com

$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.

{{< pagebreak >}}

# Part 1 Project Idea

For this project I will utilize data from the National Center for Biotechnology Information's Pathogen Detection system. This is a database that collects genomic data from a various different sources such as state laboratories, private laboratories, and government agencies like CDC, FDA, USDA, and EPA. It is primarily used a surveillance tool to monitor isolates,clusters, antimicrobial resistance, and active outbreaks. As someone that keeps track of foodborne pathogen outbreaks, I have noticed the increase in Listeria monocytogenes outbreaks. With the NCBI data I would like to visualize the occurence of new listeria strains, compare trees for clinical vs nonclincal isolates, and maybe see the correlation between the number of clinical isolates to the number of non-clincal isolates, maybe investigate how often non-clinical isolates have a clinical in the same cluster. I could also look at distribution across the United States. There is some data for other countries, but far less than the U.S. . For this project I will filter the data to show isolates collected from January 1, 2015 to January 1, 2025. This gives me 22,980 isolates and 2,365 clusters.

{{< pagebreak >}}

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

*Provide enough background on your topic that others can understand the why and how of your analysis*

## Description of data and data source

*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*

## Questions/Hypotheses to be addressed

*State the research questions you plan to answer with this analysis.*

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-03-17 code/processing-code/processingfile-v1.qmd
Part 2

---
title: "An example cleaning script"
author: "Andreas Handel"
date: "2024-02-07"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
data_location <- here::here("data","raw-data","lmonoisolates.xls")
rawdata <- readxl::read_excel(data_location)
```

# Check data

```{r}
summary(rawdata)
```

Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

```{r}
unique(rawdata$BioSample)
```
```{r}
unique(rawdata$Location)
```
```{r}
unique(rawdata$`Create date`)
```

# Cleaning

After inspecting the data,I have dedcided to remove some variables that I will not use. This will be serovar, assembly, biosample for now. I want to try and make a map so i will keep in isolate names and snp distances and what not.

```{r}
lmono2 <- rawdata %>% 
  select(-c(Serovar, Assembly, BioSample))
```
Next I want to reformat the dates so that it removes the timestamp. I will just be keeping the year, month and day
```{r}
lmono2$`Create date` <- as.Date(lmono2$`Create date`)
print(lmono2$`Create date`)
```
Next I will see the range of dates and select a 15 year time frame.
```{r}
lmono3 <- lmono2 %>%
  filter(`Create date` >= as.Date("2010-01-01") & `Create date` <= as.Date("2024-12-31"))
```
Finally I will check for any missing entries, if there are none I am good to move on.
```{r}
any(is.na(lmono3))
lmono3[!complete.cases(lmono3), ]  # Shows me what rows have missing data

```
#So this showed me that I need to remove another variable, all of the values for computed type and ...18 are NA!
```{r}
lmono4 <- lmono3 %>%
  select(-c(`Computed types`,...18))
```
#Now lets see where the missing data is 
```{r}
lmono4[!complete.cases(lmono4), ] 
```
#Now, I really want to compare food vs clinical isolates so I will make 2 new variables, One variable called Food amd one called Clinical. Food will have all entries that are food sources
```{r}
#first I am going to look at the entries, some entries are specific like hotdog or blood and some just say food or clinical, I want to include those specific ones as well so I need to see what I am working with. 
unique(lmono4$`Isolation type`)
#Some of the products under environmental other are food products so I will put environmnetal other into its own category and then filter out the non-food products since it is mostly foods.
```

```{r}
#This will create new variable "clinical"
lmono5 <- lmono4 %>%
  mutate(clinical = ifelse(`Isolation type` == "clinical", "Yes", "No"))
```

```{r}
lmono6 <- lmono5 %>%
  mutate(food = case_when(
    `Isolation type` %in% c("environmental/other", "Not Sliced-Pork", "Sliced-Pork", 
                            "Not Sliced-Turkey", "Sausage Products-Pork", 
                            "Salad/Spread/Pate-Pork", "Salad/Spread/Pate-Chicken", 
                            "Meat/Nonmeat Combination-Beef", "Meat/Nonmeat Combination", 
                            "Sliced-Beef", "Sliced-Turkey", "Not Sliced-Chicken", 
                            "Meat/Nonmeat Combination-Other Poultry", 
                            "Meat/Nonmeat Combination-Combination species", "raw", 
                            "Meat/Nonmeat Combination-Chicken", 
                            "Sausage Products-Unknown species", 
                            "Meat/Nonmeat Combination-Turkey", 
                            "Sausage Products-Beef") & 
    !`Isolation source` %in% c("environmental swab", "environmental swab sponge floor", 
                               "condensate tray", "squeegee", "chain", "wrench", 
                               "underhopper", "environmental swab sponge non food contact surface", 
                               "environmental swab sponge pallet in cooler", 
                               "environmental swab of shelf", 
                               "Environmental Isolate, sponge swab") ~ "Yes",
    TRUE ~ "No"
  ))

```
#Finally, I want to remove the observations that are stool samples that will be bovine, pigeon and catbird
```{r}
lmono_final <- lmono6 %>%
  filter(!`Isolation type` %in% c("bos taurus", "pigeon", "catbird"))
```

```{r}
processeddata <-lmono_final
```

# Save data

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location <- here::here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2025-03-17 products/manuscript/Manuscript.qmd
Part 2

@@
 
--   First Autor$^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
+-   Alexis Gonzalez $^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
 -   Second Author$^{3, *}$
--   Third Autor$^{3, \dagger}$
--   Last Author$^{1, 2, \land}$
+-   $^{3, \dagger}$
 
@@
 
-*Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*
+This project will utilize two data sources
+
+1.  National Center for Biotechnology Information's Pathogen Detection system
+2.  CDC National Notifiable Disease Surveillance System
 
@@
 
-*State the research questions you plan to answer with this analysis.*
+1.  How has the presence of listeria in food products changes over time (NCBI data)
+2.  How has the rate of listeriosis cases changed over time
+3.  How does the positive rate of food cases compare to the rate of listeriosis cases
+4.  Comparing and contrasting the isolates found in food to the isolates found in clinical samples.
+5.  What are some commonalities between establishments and locations where listeria is found and where cases are diagnosed
+6.  Follow up to 4, are these large scale companies or small scale companies.
+7.  Maybe a prediction model based to predict listeriosis cases based on number of positive isolates found?
 
@@
 
-## Data aquisition
-
-*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
-
 ## Data import and cleaning
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
+Before important the NCBI data, I filtered the data to what I wanted using the sites filters.
+I filtered data down to organism = Listeria monocytogenes, location = USA , this left me with 33,264 isolates.
+
+I will briefly use the NNDSS data because unfortunately, it is only availble from 2016 to 2022.
+The NCBI data containes clinical isolates that date farther back to we will consider those isolates to be "cases".
+Those cases are what we will compare to the source isolates (environmental, food, etc)
+
+The data cleaning code can be found in the processingfile under data-\>rawdata.
+The cleaning processed involved removing variables that I would not be using such as biosample numbers, assembly numbers, and variables that contained no data.
+Next I made some changes to date entries.
+They originally contained the time the isolate was uploaded so I removed that.
+I also created 2 new variables showing if the isolate was from a food sample or a clinical sample .
+The variable "Clinical" has a yes or no entry for if the sample is clinical, the same goes for the Food variable.
 
@@
 ## Exploratory/Descriptive analysis
+
+The full code for the EDA is under code-\>eda-code-\>eda.qmd
 
@@
 # References
+
+```{r}
+
+```

### 2025-03-18 code/eda-code/eda.qmd
Part 3

---
title: exploratory analysis script"
date: "2024-02-07"
output: html_document
---

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) 
library(dplyr)
library(skimr)
library(ggplot2)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","processeddata.rds")
#load data
mydata <- readRDS(data_location)
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

#Making a table for the clinical variable 
```{r}
table(processeddata$clinical)
```
#doing the same for food
```{r}
table(processeddata$food)
```
#Now lets visualize these tables using bar graphs
```{r}
ggplot(mydata, aes(x = clinical)) + 
  geom_bar(fill = "steelblue") + 
  labs(title = "Yes/No Distribution", x = "Response", y = "Count") + 
  theme_minimal()

```
```{r}
library(ggplot2)

ggplot(mydata, aes(x = food)) + 
  geom_bar(fill = "orange") + 
  labs(title = "Yes/No Distribution", x = "Response", y = "Count") + 
  theme_minimal()

```
#cross tabulation
```{r}
table(mydata$food, mydata$clinical)
```
#We can also look at some distribution across the country since we have some state data
```{r}

mydata %>%
  filter(food == "Yes") %>%  # Keep only "Yes" cases
  count(Location) %>%  # Count occurrences per location
  ggplot(aes(x = Location, y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Count of food isolated listeria Cases Across Locations", 
       x = "Location", 
       y = "Count of Yes") + 
  theme_minimal() +
  theme(axis.text.x = element_text(size = 6,angle = 45, hjust = 1))  # Rotate x-axis labels for readability

```
#The left most bar is USA as a whole as some entries do not report the specific state they are from, the rest are states and territories. We see that NY has more cases than other states

#Some entries show the isolation source so lets make a boxplot showing the sources for some clinical isolates.
```{r}
library(ggplot2)
library(dplyr)

mydata %>%
  dplyr::filter(clinical == "Yes") %>%  
  count(`Isolation source`) %>%
  ggplot(aes(x = `Isolation source`, y = n)) +
  geom_boxplot(fill = "darkblue") +
  labs(title = "Count of Clinical 'Yes' Across Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 4, angle = 45, hjust = 1))

#We see there are a ton of sources so we I will filter it to the top 10.
```

```{r}
mydata %>% 
  dplyr::filter(clinical == "Yes") %>%
  count(`Isolation source`) %>%
  top_n(10, n) %>%  # Keep only the top 10 sources
  ggplot(aes(x = reorder(`Isolation source`, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Top 10 Clinical 'Yes' Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))

```
#Not sure how pork ended up in the clinical category...
## Now I will try the same thing on food products

```{r}
mydata %>% 
  dplyr::filter(food == "Yes") %>%
  count(`Isolation source`) %>%
  top_n(10, n) %>%  # Keep only the top 10 sources
  ggplot(aes(x = reorder(`Isolation source`, -n), y = n)) +
  geom_bar(stat = "identity", fill = "darkblue") +
  labs(title = "Top 10 Food 'Yes' Isolation Sources", 
       x = "Isolation Source", 
       y = "Count of Yes") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1))
```
#Seeing a few issues with my data cleaning, I will go back to processing file and fix it up

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2025-03-18 code/processing-code/processingfile-v1.qmd
Part 3

@@
                                "environmental swab of shelf", 
-                               "Environmental Isolate, sponge swab") ~ "Yes",
+                               "Environmental Isolate, sponge swab","environmental swab sponge") ~ "Yes",
     TRUE ~ "No"
@@
 lmono_final <- lmono6 %>%
-  filter(!`Isolation type` %in% c("bos taurus", "pigeon", "catbird"))
+  filter(!`Isolation type` %in% c("Bos taurus\"\"", "pigeon", "catbird"))
 ```

### 2025-03-18 code/eda-code/eda.qmd
changes

@@
 ```
-#Seeing a few issues with my data cleaning, I will go back to processing file and fix it up
+#We see here that the most food isolates for listeria monocytogenes are found in Ready To Eat products, that is a term used by Food Safety and Inspection Service to describe any food products that do not require cooking or sterilization to safely consume.
 

### 2025-03-18 code/processing-code/processingfile-v1.qmd
changes

@@
                                "environmental swab of shelf", 
-                               "Environmental Isolate, sponge swab","environmental swab sponge") ~ "Yes",
+                               "Environmental Isolate, sponge swab","environmental swab sponge","Environmental-factory","Environmental-Wheels on trolley for spinner basket","Environmental-Drain basin between retort 1 and 2","bottom of sliding door in pasta room to blast","bottom of sliding door in pasta room to blast","Environmental-Bottom of sliding door in pasta room to blast","
+environmental swab from drain","environmental", "water sediment","environmental: non-food-contact surface", "soil","manure", "Environmental: food-contact surface","environment","swab") ~ "Yes",
     TRUE ~ "No"

### 2025-03-20 code/analysis-code/statistical-analysis.qmd
feedback updates

---
title: 
author: 
date: 
output: html_document
---

#Here I will perform some basic statistical analysis.

#Lets start with a basic glm of time and food and time and clinical. We will also do some visualizations using a histrogram

### 2025-03-20 code/eda-code/eda.qmd
feedback updates

@@
 ```
-
-We are saving the results to the `results` folder.
-Depending on how many tables/figures you have, it might make sense to have separate folders for each.
-And/or you could have separate folders for exploratory tables/figures and for final tables/figures.
-Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.
 
@@
 #We see here that the most food isolates for listeria monocytogenes are found in Ready To Eat products, that is a term used by Food Safety and Inspection Service to describe any food products that do not require cooking or sterilization to safely consume.
-
-# Notes
-
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.

### 2025-03-20 products/manuscript/Manuscript.qmd
feedback updates

@@
 ---
-title: "Exploring clinical and non-clinical Listeria monocytogenes isolates "
+title: "Exploring clinical and non-clinical(food) Listeria monocytogenes isolates "
 author: "Alexis Gonzalez"
@@
 
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
-
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
-
 ```{r, echo=FALSE, message=FALSE}
@@
 -   Alexis Gonzalez $^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
--   Second Author$^{3, *}$
--   $^{3, \dagger}$
 
@@
 1.  College of Public Health, University of Georgia, Athens, GA, USA.
-2.  Another Affiliation, Athens, GA, USA.
-3.  Yet another Affiliation.
-
-$*$ These authors contributed equally to this work.
-
-$\land$ Corresponding author: some\@email.com
-
-$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
 
@@
 # Summary/Abstract
-
-*Write a summary of your project.*
 
@@
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
-
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
-
 {{< pagebreak >}}
@@
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
-
 ## Schematic of workflow
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
+```{}
 ```
@@
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
-
 ## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
 
@@
 ## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
 
@@
 
-*Summarize what you did, what you found and what it means.*
-
 ## Strengths and Limitations
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
 ## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
 

### 2025-03-24 code/analysis-code/statistical-analysis.qmd
more statistical analysis

@@
 ---
+# loading in packages
+```{r}
+library(tidyverse)
+library(here) 
+library(dplyr)
+library(skimr)
+library(ggplot2)
+library(tidymodels)
+```
+#Load in the data
+```{r}
+#Path to data. 
+data_location <- here::here("data","processed-data","processeddata.rds")
+#load data
+mydata <- readRDS(data_location)
+```
 
@@
 #Lets start with a basic glm of time and food and time and clinical. We will also do some visualizations using a histrogram
+
+#To do this I have to convert my date variable into a numeric variable.
+
+```{r}
+# Load necessary package
+library(lubridate)
+
+"# Sample data
+mydata <- data.frame(
+  date `Create date` = as.Date,
+  outcome = clinical  # Yes = 1, No = 0
+)"
+
+# Recode Yes = 1, No = 0
+mydata$clinical_numeric <- ifelse(mydata$clinical == "Yes", 1, 0)
+
+# Convert date to numeric (days since first observation)
+mydata$days_since_start <- as.numeric(difftime(mydata$`Create date`, min(mydata$`Create date`), units = "days"))
+
+# Logistic regression
+model <- glm(clinical_numeric ~ days_since_start, data = mydata, family = binomial)
+
+# Model summary
+summary(model)
+
+```
+```{r}
+
+ggplot(mydata, aes(x = days_since_start, y = clinical_numeric)) +
+  geom_jitter(height = 0.05, width = 5, alpha = 0.5) +  
+  geom_line(data = new_data, aes(x = days_since_start, y = predicted_prob), color = "blue", size = 1) +
+  labs(title = "Logistic Regression: Probability of Clinical Isolates Over Time",
+       x = "Days Since Start",
+       y = "Probability of Outcome (Yes = 1)") +
+  theme_minimal()
+
+```
+```{r}
+ggplot(mydata, aes(x = days_since_start, fill = factor(clinical_numeric))) +  # Fill by outcome (0 or 1)
+  geom_histogram(position = "identity", 
+                 alpha = 0.5, 
+                 bins = 30, 
+                 color = "black") +  # Standard histogram
+  labs(title = "Histogram of Outcomes Over Time",
+       x = "Days Since Start",
+       y = "Count of Outcome (0 or 1)",
+       fill = "Outcome") +
+  theme_minimal()
+
+```

### 2025-03-24 code/processing-code/processingfile-v1.qmd
more statistical analysis

@@
 ---
-title: "An example cleaning script"
-author: "Andreas Handel"
+title: " cleaning script"
+author: "Alexis Gonzalez"
 date: "2024-02-07"
@@
 ---
-
-# Processing script
-
-This contains the same code and comments/information as `processingcode.R`.
-
-This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.
-
-See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.
 
@@
 ```{r}
-library(readxl) #for loading Excel files
-library(dplyr) #for data processing/cleaning
-library(tidyr) #for data processing/cleaning
-library(skimr) #for nice visualization of data 
-library(here) #to set paths
+library(readxl) 
+library(dplyr) 
+library(tidyr)
+library(skimr) 
+library(here) 
 ```
@@
 ```
-Finally I will check for any missing entries, if there are none I am good to move on.
+Finally I will check for any missing entries
 ```{r}

### 2025-03-24 products/manuscript/Manuscript.qmd
more statistical analysis

@@
 
+Will complete at the end
+
 {{< pagebreak >}}
@@
 
-*Provide enough background on your topic that others can understand the why and how of your analysis*
+*Listeria monocytogenes* is a ubiquitous, gram positive, intra-cellular pathogen renowned for its adaptability and resilience (cite).
+This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation.
+Due to its ability to proliferate in food products and low temperatures, it is considered a major food-borne illness.
+The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death.
+It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[cite\].
+Listeriosis causes 1,600 infections and 260 deaths each year with an over 95% hospitalization rate.
+Listeria can be found in a wide range of meats, produce, and dairy, such as deli meats, soft cheese, and melons.
+The Centers for Disease Control and Prevention has deemed *Listeria monocytogenes* as a nationally notifiable disease which is a disease that federal agencies, healthcare providers, and laboratories are required to report.
+All positive isolate data found in the environment, food samples, clinical samples and animal samples are uploaded to the National Center for Biotechnology Information's Pathogen Detection system.
 
@@
 
-```{}
+```         
 ```
@@
 
-I will briefly use the NNDSS data because unfortunately, it is only availble from 2016 to 2022.
-The NCBI data containes clinical isolates that date farther back to we will consider those isolates to be "cases".
+I will briefly use the NNDSS data because unfortunately, it is only available from 2016 to 2022.
+The NCBI data contains clinical isolates that date farther back to we will consider those isolates to be "cases".
 Those cases are what we will compare to the source isolates (environmental, food, etc)

### 2025-04-05 products/manuscript/Manuscript.qmd
manuscript updates

@@
 
-# Part 1 Project Idea
-
-For this project I will utilize data from the National Center for Biotechnology Information's Pathogen Detection system.
-This is a database that collects genomic data from a various different sources such as state laboratories, private laboratories, and government agencies like CDC, FDA, USDA, and EPA.
-It is primarily used a surveillance tool to monitor isolates,clusters, antimicrobial resistance, and active outbreaks.
-As someone that keeps track of foodborne pathogen outbreaks, I have noticed the increase in Listeria monocytogenes outbreaks.
-With the NCBI data I would like to visualize the occurence of new listeria strains, compare trees for clinical vs nonclincal isolates, and maybe see the correlation between the number of clinical isolates to the number of non-clincal isolates, maybe investigate how often non-clinical isolates have a clinical in the same cluster.
-I could also look at distribution across the United States.
-There is some data for other countries, but far less than the U.S.
-.
-For this project I will filter the data to show isolates collected from January 1, 2015 to January 1, 2025.
-This gives me 22,980 isolates and 2,365 clusters.
-
-{{< pagebreak >}}
-
 # Summary/Abstract
@@
 
-# Introduction
+#  1. Introduction
 
-## General Background Information
-
-*Listeria monocytogenes* is a ubiquitous, gram positive, intra-cellular pathogen renowned for its adaptability and resilience (cite).
+Listeria monocytogenes is a ubiquitous, gram-positive, intracellular pathogen renowned for its adaptability and resilience\[5\].
 This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation.
@@
 The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death.
-It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[cite\].
-Listeriosis causes 1,600 infections and 260 deaths each year with an over 95% hospitalization rate.
-Listeria can be found in a wide range of meats, produce, and dairy, such as deli meats, soft cheese, and melons.
-The Centers for Disease Control and Prevention has deemed *Listeria monocytogenes* as a nationally notifiable disease which is a disease that federal agencies, healthcare providers, and laboratories are required to report.
-All positive isolate data found in the environment, food samples, clinical samples and animal samples are uploaded to the National Center for Biotechnology Information's Pathogen Detection system.
+It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[10\].
+While immunocompetent individuals may only experience mild gastroenteritis, disseminated infection in the immunocompromised has a lethality rate of \~30%\[8\].  Listeriosis causes 1,600 infections and 260 deaths each year with a 95% hospitalization rate.
+The United States Department of Agriculture lists 6 foods that are at high risks for L.monocytogenes contamination; unpasteurized dairy products, soft cheeses, raw fruits and vegetables (particularly sprouts, enoki mushrooms, and melons), Ready-To-Eat deli meats and hot dogs, refrigerated pate, and refrigerated smoked seafood.
+Listeria contamination typically happens post-lethality which exposes a ready-to-eat product to a processing environment after a form of lethality treatment has taken place.
+Post lethality processing environments can be deli slicers, submersion in a brine solution, meat encassing, bagging or packaging.
+The USDA recommends all establishments to take part in additional post lethality treatment such as antimicrobial coatings, pasteurization, or chemical or physical sanitation of equipment.
+Sanitation tends to be the more cost effective option for many establishments but, due to incorrect sanitation practices and L.monocytogenes robust bio-film formation capabilities sanitation does not always prevent contamination.
+Outbreaks have been seen in a range of products in recent years.
+From 2023 to 2025 there have been over 10 multi-state listeria outbreaks.
+The most recent being an outbreak beginning February 22, 2025 in frozen supplement shakes.
+These shakes were distributed among institutional facilities such as long term care facilities and hospitals.
+This outbreak has resulted in 38 cases, 37 hospitalizations, and 12 deaths so far as this investigation is ongoing as of April 2025.
 
-## Description of data and data source
-
-This project will utilize two data sources
-
-1.  National Center for Biotechnology Information's Pathogen Detection system
-2.  CDC National Notifiable Disease Surveillance System
-
-## Questions/Hypotheses to be addressed
-
-1.  How has the presence of listeria in food products changes over time (NCBI data)
-2.  How has the rate of listeriosis cases changed over time
-3.  How does the positive rate of food cases compare to the rate of listeriosis cases
-4.  Comparing and contrasting the isolates found in food to the isolates found in clinical samples.
-5.  What are some commonalities between establishments and locations where listeria is found and where cases are diagnosed
-6.  Follow up to 4, are these large scale companies or small scale companies.
-7.  Maybe a prediction model based to predict listeriosis cases based on number of positive isolates found?
+\
+This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of a year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products.
+The data source being used to conduct this study is the National Center for Biotechnology Information\'s Pathogen Detection browser.
+Pathogen Detection is a culmination of pathogen genomic sequences from federal, state, private, and foreign laboratories.
+It is utilized constantly by public health officials,epidemiologists, and scientists to identify chains of transmission.
+The new isolates are analyzed in real time to aid in that identification by developing clusters.
+Listeria monocytogenes is a nationally notifiable disease so, many laboratories opt to share their data nationally through the Pathogen Detection Browser.
+By utilizing the most up to date data and identifying trends, this study hopes to provide information to form new guidelines on protecting the public from Listeriosis infection.
 
@@
 
-# Methods
+# 2. Methods
 
-## Schematic of workflow
+##
 
@@
 
-Before important the NCBI data, I filtered the data to what I wanted using the sites filters.
-I filtered data down to organism = Listeria monocytogenes, location = USA , this left me with 33,264 isolates.
-
-I will briefly use the NNDSS data because unfortunately, it is only available from 2016 to 2022.
-The NCBI data contains clinical isolates that date farther back to we will consider those isolates to be "cases".
-Those cases are what we will compare to the source isolates (environmental, food, etc)
-
-The data cleaning code can be found in the processingfile under data-\>rawdata.
-The cleaning processed involved removing variables that I would not be using such as biosample numbers, assembly numbers, and variables that contained no data.
-Next I made some changes to date entries.
-They originally contained the time the isolate was uploaded so I removed that.
-I also created 2 new variables showing if the isolate was from a food sample or a clinical sample .
-The variable "Clinical" has a yes or no entry for if the sample is clinical, the same goes for the Food variable.
+This study utilized publicly available data through NCBI Pathogen Detection.
+Before downloading the data from the Isolate Browser, the data was filtered to only include Listeria monocytogenes isolates found in or originating in the United States of America.
+This resulted in 33,264 L.monocytogenes isolates.
+Once loaded into R the data set was minimized further by selecting isolates that were entered between January 1, 2010 and December 31, 2024.
+Several variables were removed from the dataset as they did not contribute to the analysis.
+These were isolate identification and clustering data such as Serovar, Biosample number, Assembly, Min-same and Min-diff SNPs, computed types, and a variable called ...18 ,presumably referring to 18s primers, that contained no data.
 
@@
 ```
+
+*Listeria monocytogenes* is a ubiquitous, gram positive, intra-cellular pathogen renowned for its adaptability and resilience (cite).
+This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation.
+Due to its ability to proliferate in food products and low temperatures, it is considered a major food-borne illness.
+The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death.
+It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[cite\].Listeriosis causes 1,600 infections and 260 deaths each year with an over 95% hospitalization rate.
+Listeria can be found in a wide range of meats, produce, and dairy, such as deli meats, soft cheese, and melons.
+The Centers for Disease Control and Prevention has deemed *Listeria monocytogenes* as a nationally notifiable disease which is a disease that federal agencies, healthcare providers, and laboratories are required to report.
+All positive isolate data found in the environment, food samples, clinical samples and animal samples are uploaded to the National Center for Biotechnology Information's Pathogen Detection system.

### 2025-04-06 code/analysis-code/statistical-analysis.qmd
adding some figures to manuscript

@@
   geom_jitter(height = 0.05, width = 5, alpha = 0.5) +  
-  geom_line(data = new_data, aes(x = days_since_start, y = predicted_prob), color = "blue", size = 1) +
+  geom_line(data = mydata, aes(x = days_since_start, y = predicted_prob), color = "blue", size = 1) +
   labs(title = "Logistic Regression: Probability of Clinical Isolates Over Time",
@@
 ```{r}
-ggplot(mydata, aes(x = days_since_start, fill = factor(clinical_numeric))) +  # Fill by outcome (0 or 1)
+p <- ggplot(mydata, aes(x = days_since_start, fill = factor(clinical_numeric))) +  # Fill by outcome (0 or 1)
   geom_histogram(position = "identity", 
@@
                  color = "black") +  # Standard histogram
-  labs(title = "Histogram of Outcomes Over Time",
+  labs(title = "Histogram of clinical isolates Over Time",
        x = "Days Since Start",
@@
 
+figure_file=here("results/figures","clinicalxtime.png")
+ggsave(filename=figure_file, plot = p, width = 8, height=6)
 ```
+
+```

### 2025-04-06 code/eda-code/eda.qmd
adding some figures to manuscript

@@
 ```{r}
-ggplot(mydata, aes(x = clinical)) + 
+bar1 <- ggplot(mydata, aes(x = clinical)) + 
   geom_bar(fill = "steelblue") + 
@@
 
-ggplot(mydata, aes(x = food)) + 
+p1 <- ggplot(mydata, aes(x = food)) + 
   geom_bar(fill = "orange") + 
@@
   theme_minimal()
+figure_file=here("results/figures","foodisolatecount.png")
+ggsave(filename=figure_file, plot = p1, width = 8, height=6)
 

### 2025-04-06 products/manuscript/Manuscript.qmd
adding some figures to manuscript

@@
 This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of a year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products.
-The data source being used to conduct this study is the National Center for Biotechnology Information\'s Pathogen Detection browser.
+The data source being used to conduct this study is the National Center for Biotechnology Information's Pathogen Detection browser.
 Pathogen Detection is a culmination of pathogen genomic sequences from federal, state, private, and foreign laboratories.
@@
 
-##
-
-```         
-```
-
-## Data import and cleaning
-
 This study utilized publicly available data through NCBI Pathogen Detection.
@@
 
-## Statistical analysis
-
-*Explain anything related to your statistical analyses.*
-
 {{< pagebreak >}}
@@
 
-The full code for the EDA is under code-\>eda-code-\>eda.qmd
+```{r}
+knitr::include_graphics(here("results/figures", "foodisolatecount.png"))
+```
+
+```{r}
+
+```
 
@@
 ```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
+
 ```
@@
 
+#Need to insert figures
+
+```{r}
+knitr::include_graphics(here("results/figures", "clinicalxtime.png"))
+
+```
+
 ```{r}
 #| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
+#| tbl-cap: "Linear model fit table."...................;;;;;;;;;;;;;;;::::::::::::::::P
 #| echo: FALSE
@@
 ```
-
-*Listeria monocytogenes* is a ubiquitous, gram positive, intra-cellular pathogen renowned for its adaptability and resilience (cite).
-This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation.
-Due to its ability to proliferate in food products and low temperatures, it is considered a major food-borne illness.
-The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death.
-It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[cite\].Listeriosis causes 1,600 infections and 260 deaths each year with an over 95% hospitalization rate.
-Listeria can be found in a wide range of meats, produce, and dairy, such as deli meats, soft cheese, and melons.
-The Centers for Disease Control and Prevention has deemed *Listeria monocytogenes* as a nationally notifiable disease which is a disease that federal agencies, healthcare providers, and laboratories are required to report.
-All positive isolate data found in the environment, food samples, clinical samples and animal samples are uploaded to the National Center for Biotechnology Information's Pathogen Detection system.

### 2025-04-07 code/analysis-code/statistical-analysis.qmd
additions to statistical analysis

@@
 library(tidymodels)
+library(lubridate)
+library(ranger)
+
 ```
@@
 
+#More advanced temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
+#For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.
+
+```{r}
+mydata <- mydata %>%
+  drop_na()%>%
+  mutate(
+    `Create date` = as.Date(`Create date`),
+    year = year(`Create date`),
+    day_of_year= yday(`Create date`),
+    week = week(`Create date`),
+    month = month(`Create date`),
+    season = case_when(
+      month %in% c(12, 1, 2)  ~ "Winter",
+      month %in% c(3, 4, 5)   ~ "Spring",
+      month %in% c(6, 7, 8)   ~ "Summer",
+      month %in% c(9, 10, 11) ~ "Fall"
+    ),
+    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
+    `Isolation source` = factor(`Isolation source`)
+  )
 ```
+
+#Feature selection
+```{r}
+model1_data <- mydata %>%
+  select(`Isolation type`,year,month,day_of_year,week,season)
+```
+
+#Split the data
+```{r}
+set.seed(1234)
+data_split <- initial_split(model1_data, prop = 0.8, strata = `Isolation type`)
+train_data <-training(data_split)
+test_data <-testing(data_split)
+```
+#Create recipe
+```{r}
+rf_recipe <- recipe(`Isolation type` ~ month + year + day_of_year + week + season, data = train_data) %>%
+  step_dummy(all_nominal_predictors()) %>%
+  step_zv(all_predictors()) %>%
+  step_normalize(all_numeric_predictors())
+```
+
+#Define the model
+```{r}
+rf_mod <-rand_forest(trees = 150, min_n =5) %>%
+  set_mode("classification") %>%
+  set_engine("ranger")
+```
+Create workflow
+```{r}
+workflow1 <- workflow() %>%
+  add_model(rf_mod)%>%
+  add_recipe(rf_recipe)
+```
+#fit the random forest model
+```{r}
+fit1 <- workflow1 %>%
+  fit(data = train_data)
+```
+#Make predictions for evaluation
+```{r}
+predictions_rf <-fit1 %>%
+  predict(new_data = test_data) %>%
+  bind_cols(test_data)
+
+```
+```{r}
+#set isolation source as a factor
+predictions_rf <- predictions_rf %>%
+  mutate(`Isolation type` = as.factor(`Isolation type`))
+
+```
+
+#Run metrics to evaluate the model
+```{r}
+
+# Create a metric set for classification
+classification_metrics <- metric_set(accuracy, precision, recall, f_meas)
+
+# Now use it on your predictions
+metrics_rf <- predictions_rf %>%
+  classification_metrics(truth = `Isolation type`, estimate = .pred_class)
+
+```
+
+```{r}
+metrics_rf <- predictions_rf %>%
+  metrics(truth = `Isolation type`, estimate = .pred_class)
+print(metrics_rf)
+```
+#Plot the results of this model
+#Clean up some of the changes made to the data and add to processing file

### 2025-04-20 project-review-rayleen-lewis.qmd
Part review - Rayleen Lewis

---
title: Project Review Template 
author: Rayleen Lewis
date: April 20, 2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Exploring clinical and non-clinical(food) Listeria monocytogenes isolates

Name of project author(s): Alexis Gonzalez

Name of project reviewer: Rayleen Lewis

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/renders into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

# Specific project content evaluation
Evaluate the different parts of the project by filling in the sections below.

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Information on the burden and sources of Listeria outbreaks is provided, as is information on recent Listeria outbreaks. The information included is good, but the intro doesn't have any citations. These need to be added for the final project.

### Summary assessment (PICK ONE, DELETE THE OTHERS)]
* some contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

The objective is somewhat clear. Adding definitions of "clinical" and "non-clinical" samples would be helpful. Are clinical from patients seeking medical treatment while non-clinical are directly from food/food processing equipment? The data seem appropriate for answering the question. There isn't a hypothesis, but this seems more of an exploratory aim, so I think that's fine.

### Summary assessment
* question/hypotheses somewhat explained

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

The overall purpose of the data source is described. It is unclear what variables/information are included in the data; only variables that are excluded were mentioned. References should be added for the data sources. This could be a reference to a website if data are available online or a methods paper describing the platforms.

### Summary assessment
* source and overall structure of data poorly explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

Not all characteristics were looked at during the exploratory data analysis, but I'm not sure if this was appropriate since a description of the characteristics wasn't included. For example, would it have been useful to look at cluster and see if the case counts per cluster made sense? Also, did you consider recategorizing location to be state level? Most of the locations are state, but some are city. I'm not sure if this would be of interest in your analysis. There is no supplementary material. Wrangling steps are described in the methods.

### Summary assessment
* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

There's virtually no analysis in the manuscript. The manuscript includes 2 figures, which don't have descriptive titles or legends. Please explain what Yes/No response is in reference to for the first figure, and 0/1 for outcome and starting point for days counter in the second figure. For the analysis code itself, if the goal is to look at trends, I don't think it's appropriate to use day of the year, week, and month as separate predictors in your model. These are essentially providing the same information at different levels of detail. For example, weeks 1-4 will all have a month = 1.

### Summary assessment
* wrong/inadequate analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

See comments above about improving figures. No other analyses are included in the manuscript. Also, code to execute inclusion of the figures should be removed. The manuscript file also includes a leftover table from the template.

### Summary assessment
* results are poorly presented, hard to understand, poor quality

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

This section is missing entirely.

### Summary assessment
* major parts of discussion missing or wrong

## Further comments

Because of the major gaps in the project, I'm wondering if updates on the project weren't pushed to GitHub.

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Most of the template files are still in the folders, so it is confusing to figure out which files were leftover from the template and which files were relevant for this project. The file names are all from the template, so they make sense. Files do appear in the expected folders (e.g., figures are in the figures folder).

### Summary assessment
* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

All files include some level of comments, including some text explaining findings from figures. For some code blocks this is minimal. For example, some comments are very broad like "build model" and "run metrics." More details, like what type of model is being built or which metrics you're using to evaluate model performance, would be helpful.

### Summary assessment
* decently documented with some gaps

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments
All of the ReadMe files should be updated to describe this project and should include information on what code should be run in what order. Because a lot of files are left over from the template, it is unclear which files I should run to reproduce the project.

I did get some errors when running the code files that I think are for this project and not the template:
Statistical analysis code:
- Lines 60-66:  I couldn't reproduce this figure (error: object 'predicted_prob' not found)

Manuscript code:
- Line 79: This was added to the end of the line which prevented the document from rendering "...................;;;;;;;;;;;;;;;::::::::::::::::P"

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

Most of the analysis is missing from the manuscript, the statistical code only ran a single model (random forest), and the discussion is not included.

### Summary assessment
* weak level of thoroughness

## Further comments

It's very important to update the ReadMe files so someone new to the project knows what to expect in the folders and which files to run. Also, deleting template files that aren't part of the project would help a lot when this is reviewed for final grading.

### 2025-04-21 code/analysis-code/statistical-analysis.qmd
updates to manuscript

@@
 library(ranger)
-
+library(parsnip)
+library(poissonreg)
+library(randomForest)
 ```
@@
 ```
-```{r}
 
-ggplot(mydata, aes(x = days_since_start, y = clinical_numeric)) +
-  geom_jitter(height = 0.05, width = 5, alpha = 0.5) +  
-  geom_line(data = mydata, aes(x = days_since_start, y = predicted_prob), color = "blue", size = 1) +
-  labs(title = "Logistic Regression: Probability of Clinical Isolates Over Time",
-       x = "Days Since Start",
-       y = "Probability of Outcome (Yes = 1)") +
-  theme_minimal()
-
-```
-```{r}
-p <- ggplot(mydata, aes(x = days_since_start, fill = factor(clinical_numeric))) +  # Fill by outcome (0 or 1)
-  geom_histogram(position = "identity", 
-                 alpha = 0.5, 
-                 bins = 30, 
-                 color = "black") +  # Standard histogram
-  labs(title = "Histogram of clinical isolates Over Time",
-       x = "Days Since Start",
-       y = "Count of Outcome (0 or 1)",
-       fill = "Outcome") +
-  theme_minimal()
-
-figure_file=here("results/figures","clinicalxtime.png")
-ggsave(filename=figure_file, plot = p, width = 8, height=6)
-```
-
-#More advanced temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
+# temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
 #For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.
@@
   set_mode("classification") %>%
-  set_engine("ranger")
+  set_engine("ranger",importance="impurity")
 ```
@@
 ```{r}
-
 # Create a metric set for classification
 classification_metrics <- metric_set(accuracy, precision, recall, f_meas)
+```
 
-# Now use it on your predictions
+```{r}
+#  use it on predictions
 metrics_rf <- predictions_rf %>%
@@
 #Plot the results of this model
-#Clean up some of the changes made to the data and add to processing file
+
+```{r}
+library(vip)  # for variable importance plot
+
+# Extract the fitted model from the workflow
+rf_fit <- extract_fit_parsnip(fit1)
+
+# Plot variable importance
+vip(rf_fit$fit)
+
+```
+
+#Poisson modeling using month
+
+```{r}
+# Create monthly clinical counts
+monthly_counts <- mydata %>%
+  filter(clinical == "Yes") %>%
+  mutate(month = floor_date(`Create date`, "month")) %>%
+  group_by(month) %>%
+  summarize(clinical_count = n()) %>%
+  ungroup()
+#filling in gaps with 0
+monthly_counts <- monthly_counts %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(clinical_count = 0))
+#Feature engineering, extracting the month from the data
+monthly_counts <- monthly_counts %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+```
+
+```{r}
+library(tidymodels)
+library(parsnip)
+#Splitting the data 
+set.seed(123)
+data_split2 <- initial_time_split(monthly_counts, prop = 0.8)
+train_data2 <- training(data_split2)
+test_data2  <- testing(data_split2)
+
+```
+```{r}
+#create a recipe 
+poisson_recipe <- recipe(clinical_count ~ month_name, data = train_data2) %>%
+  step_dummy(all_nominal_predictors())
+
+```
+```{r}
+#Select the model
+poisson_model <- 
+  poisson_reg(
+    mode = "regression",
+    penalty = NULL,
+    mixture = NULL,
+    engine ="glm"
+  )
+
+```
+```{r}
+#Create a workflow 
+library(poissonreg)
+poisson_wf <-workflow() %>%
+  add_model(poisson_model) %>%
+  add_recipe(poisson_recipe)
+```
+```{r}
+#fit the model
+poisson_fit <- poisson_wf %>%
+  fit(data = train_data2)
+```
+```{r}
+#make predictions
+predictions <- predict(poisson_fit, new_data = train_data2) %>%
+  bind_cols(train_data2)
+metrics(predictions, truth = clinical_count, estimate =.pred)
+```
+#Visualize results
+```{r}
+poisson_seasonal <- ggplot(predictions, aes(x = month)) +
+  geom_line(aes(y = clinical_count), color = "black", linetype = "dashed") +
+  geom_line(aes(y = .pred), color = "blue") +
+  labs(title = "Predicted vs Actual Monthly Clinical Isolate Counts",
+       y = "Clinical Count")
+print(poisson_seasonal)
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/poisson_seasonal.png", plot = poisson_seasonal, width = 8, height = 6, dpi = 300)
+```
+#Predict future isolate counts based on seasonality
+```{r}
+last_month <-max(monthly_counts$month)
+```
+
+```{r}
+future_months <- tibble(
+  month = seq.Date(from = last_month %m+% months(1),by = "month", length.out = 12)
+)%>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered= FALSE))
+```
+
+```{r}
+#make predictions
+future_predictions <- predict(poisson_fit, new_data = future_months)%>%
+  bind_cols(future_months)
+```
+```{r}
+#visualize
+# Combine historical + future for plotting
+combined_months<- bind_rows(
+  monthly_counts %>% select(month, clinical_count) %>% mutate(source = "Actual"),
+  future_predictions %>% rename(clinical_count = .pred) %>% mutate(source = "Forecast")
+)
+
+# Plot
+seasonalpreds <- ggplot(combined_data, aes(x = month, y = clinical_count, color = source)) +
+  geom_line() +
+  labs(title = "Actual and Forecasted Clinical Isolate Counts",
+       y = "Clinical Count",
+       x = "Month") +
+  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
+  theme_minimal()
+print(seasonalpreds)
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", plot = seasonalpreds, width = 8, height = 6, dpi = 300)
+```
+
+```{r}
+
+# 1. Get the last month from your historical data
+last_month <- max(monthly_counts$month)
+
+# 2. Create the next 12 months
+future_months <- tibble(
+  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
+) %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+# 3. Make predictions using your Poisson model (or other model)
+future_predictions <- predict(poisson_fit, new_data = future_months) %>%
+  bind_cols(future_months)
+
+# 4. Combine historical and forecast data
+combined_months <- bind_rows(
+  monthly_counts %>% 
+    select(month, clinical_count) %>% 
+    mutate(source = "Actual"),
+  future_predictions %>% 
+    rename(clinical_count = .pred) %>% 
+    mutate(source = "Forecast")
+)
+
+# 5. Plot
+seasonalpreds <- ggplot(combined_months, aes(x = month, y = clinical_count, color = source)) +
+  geom_line(size = 1.2) +
+  labs(
+    title = "Actual and Forecasted Clinical Isolate Counts",
+    y = "Clinical Count",
+    x = "Month"
+  ) +
+  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+# 6. Show plot
+print(seasonalpreds)
+
+# 7. Save to file
+ggsave(
+  "/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", 
+  plot = seasonalpreds, width = 8, height = 6
+
+```
+
+```{r}
+library(mgcv)
+gam_model <- gam(clinical_count ~ s(day_of_year), family = poisson(), data = daily_counts)
+plot(gam_model)
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png", plot = gam_model, width = 8, height = 6, dpi = 300)
+```
+
+```{r}
+# Load necessary library
+library(mgcv)
+library(gratia) # for tidy ggplot GAM visualizations
+
+# Example: Create ggplot-style plot from your GAM model
+gam_plot <- draw(gam_model)  # `gam_model` is your fitted model
+
+# Now save it
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png",
+       plot = gam_plot, width = 8, height = 6, dpi = 300)
+
+```
+
+```{r}
+
+#  Split the data into training and testing sets
+set.seed(123)
+data_split <- initial_split(monthly_counts, prop = 0.8)
+train_data <- training(data_split)
+test_data <- testing(data_split)
+
+#  Define the recipe (clinical_count predicted by food_count)
+poisson_recipe <- recipe(clinical_count ~ food_count, data = train_data)
+
+# Define the Poisson regression model
+poisson_model <- poisson_reg(mode = "regression", engine = "glm")
+
+#  Create a workflow 
+poisson_wf <- workflow() %>%
+  add_recipe(poisson_recipe) %>%
+  add_model(poisson_model)
+
+#  Fit the model
+poisson_fit <- poisson_wf %>%
+  fit(data = train_data)
+
+# Make predictions on the training data
+predictions <- predict(poisson_fit, new_data = train_data) %>%
+  bind_cols(train_data)
+
+#  Evaluate the model 
+isolate_metrics <- metrics(predictions, truth = clinical_count, estimate = .pred)
+print(isolate_metrics)
+
+# Visualize the results
+ggplot(predictions, aes(x = month)) +
+  geom_col(aes(y = clinical_count), fill = "lightgray", alpha = 0.8) +
+  geom_line(aes(y = .pred), color = "blue", size = 1) +
+  labs(
+    title = "Poisson Regression: Predicting Clinical Isolates from Food Isolates",
+    y = "Clinical Isolate Count",
+    x = "Month"
+  ) +
+  theme_minimal()
+```
+#Chart showing likelihood
+```{r}
+
+```
+
+#Now I want to do one more advanced statistical analysis.
+This analysis will utilize the counts of clinical and counts of food isolates each year and see if the amount of clinical isolates can be predicted by the amount of food isolates.
+This purpose of this analysis is to see if federal reguulations are doing a good job of preventing illness.
+
+#put my food and clinical variable into counts and group by year
+
+```{r}
+monthly_counts2 <- mydata %>%
+  mutate(month = floor_date(`Create date`, "month")) %>%
+  count(month, isolation_source) %>%
+  tidyr::pivot_wider(names_from = isolation_source, values_from = n, values_fill = 0)
+```
+
+```{r}
+set.seed(123)  # for reproducibility
+
+rf_model <- randomForest(
+  clinical ~ food + environmental,
+  data = monthly_counts2,
+  ntree = 500,          # number of trees
+  importance = TRUE     # track variable importance
+)
+print(rf_model)
+
+```
+
+```{r}
+#model_predict_clinical <- lm(clinical ~ food + environmental, data = monthly_counts2)
+#summary(model)
+
+```{r}
+
+```
+
+`
+#Visualzing the future
+```{r}
+# Suppose you created this:
+future_data <- data.frame(
+  month = as.Date(c("2025-02-01","2025-03-01","2025-04-01","2025-05-01", "2025-06-01", "2025-07-01","2025-08-01","2025-09-01","2025-10-01","2025-11-01","2025-12-01")),
+  food = c(300,220, 180, 250,140,122,116,235,96,175,225),
+  environmental = c(350,150, 170, 200,220,140,267,178,155,200,280)
+)
+```
+
+```{r}
+# Predict clinical isolates
+future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)
+```
+
+```{r}
+# Predict clinical isolates for future months
+future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)
+
+# Combine with past data (no predicted clinicals in the past, so we set it to NA)
+combined_plot_data <- monthly_counts2 %>%
+  select(month, clinical) %>%
+  mutate(Predicted_Clinical = NA,
+         Source = "Historical") %>%
+  bind_rows(
+    future_data %>%
+      transmute(month, clinical = NA, Predicted_Clinical, Source = "Forecast")
+  )
+
+# Filter combined_plot_data for the last 3 years
+recent_data <- combined_plot_data %>%
+  filter(month >= (max(month) %m-% years(3)))
+
+# Plot
+ggplot(recent_data, aes(x = month)) +
+  geom_line(aes(y = Predicted_Clinical, color = Source), size = 1.2, linetype = "dashed") +
+  geom_line(data = recent_data %>% filter(!is.na(clinical)),
+            aes(y = clinical, color = "Observed"), size = 1) +
+  scale_color_manual(values = c("Observed" = "steelblue", "Historical" = "tomato", "Forecast" = "forestgreen")) +
+  labs(title = "Observed and Forecasted Clinical Isolates (Past 3 Years)",
+       x = "Month",
+       y = "Count",
+       color = "Legend") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+```{r}
+# Create new data (adjust as needed)
+new_data <- data.frame(food = c(200), environmental = c(150))
+
+# Predict clinical isolates
+predict(model_predict_clinical, newdata = new_data)
+
+```
+#Machine Learning since this was just a simple linear regression
+#Random forest
+```{r}
+#set seed and split data
+set.seed(123)
+data_split3 <- initial_split(model_data, prop = 0.8)
+train_data3 <- training(data_split3)
+test_data3 <- testing(data_split3)
+
+```

### 2025-04-21 code/eda-code/eda.qmd
updates to manuscript

--- 2025-03-20 code/eda-code/eda.qmd
@@
 ```
+#Creating an additional variable called "Isolation_source" This is different from the `Isolation Source` variable in that it will include food and not just list it under environmental/other
+```{r}
+mydata <- mydata %>%
+  mutate(
+    clinical = as.character(clinical),
+    food = as.character(food),
+    isolation_source = case_when(
+      clinical == "Yes" ~ "clinical",
+      food == "Yes" ~ "food",
+      TRUE ~ "environmental"
+    )
+  )
+
+```
 
@@
 ```{r}
-table(processeddata$clinical)
+table(mydata$clinical)
 ```
@@
 ```{r}
-table(processeddata$food)
-```
-#Now lets visualize these tables using bar graphs
-```{r}
-ggplot(mydata, aes(x = clinical)) + 
-  geom_bar(fill = "steelblue") + 
-  labs(title = "Yes/No Distribution", x = "Response", y = "Count") + 
-  theme_minimal()
-
+table(mydata$food)
 ```
 ```{r}
-library(ggplot2)
+table(mydata$isolation_source)
+```
 
-ggplot(mydata, aes(x = food)) + 
-  geom_bar(fill = "orange") + 
-  labs(title = "Yes/No Distribution", x = "Response", y = "Count") + 
+#Now lets visualize these tables using bar graphs of counts for food clincial and env
+```{r}
+
+isolatetypes <- ggplot(mydata, aes(x = isolation_source)) +
+  geom_bar(fill = "seagreen") +
+  labs(title = "Isolation Source Distribution",
+       x = "Isolation Source",
+       y = "Count") +
   theme_minimal()
-
+ print(isolatetypes)  
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/isolatetypes.png", plot = isolatetypes, width = 8, height = 6, dpi = 300)
 ```
+#As we see here, clinical cases are rare but that does not take away from the fact that they are usually severe.
+```{r}
 #cross tabulation
-```{r}
 table(mydata$food, mydata$clinical)
@@
 #We can also look at some distribution across the country since we have some state data
+
 ```{r}
 
+locations <- ggplot(
   mydata %>% 
-  filter(food == "Yes") %>%  # Keep only "Yes" cases
-  count(Location) %>%  # Count occurrences per location
-  ggplot(aes(x = Location, y = n)) +
+    filter(food == "Yes", Location != "USA"),
+  aes(x = Location)
+) + 
+  geom_bar(fill = "steelblue") + 
+  labs(
+    title = "Count of Food-Isolated Listeria Cases Across Locations",
+    x = "Location",
+    y = "Count of Yes"
+  ) + 
+  theme_minimal() + 
+  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
+print(locations)
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/locations.png", plot = locations, width = 8, height = 6, dpi = 300)
+```
+
+```{r}
+locations2 <- ggplot(mydata %>% filter(clinical == "Yes"), aes(x = Location)) +
+  geom_bar(fill = "steelblue") +
+  labs(title = "Count of Clincally-Isolated Listeria Cases Across Locations",
+       x = "Location",
+       y = "Count") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
+print(locations2)
+```
+#This only tells me that all cases of listeriosis were in the U.S. which i specified for so we wont use this graph.
+
+#The left most bar is USA as a whole as some entries do not report the specific state they are from, the rest are states and territories. We see that NY has more cases than other states
+```{r}
+sourcestate <- mydata %>%
+  filter(clinical == "Yes", !is.na(`Food origin`)) %>%  
+  count(`Food origin`) %>%  
+  ggplot(aes(x = `Food origin`, y = n)) +
   geom_bar(stat = "identity", fill = "steelblue") +
-  labs(title = "Count of food isolated listeria Cases Across Locations", 
-       x = "Location", 
-       y = "Count of Yes") + 
+  labs(title = "Count of Clinically Isolated Listeria Sources by Food Origin",
+       x = "Food Origin",
+       y = "Count") +
   theme_minimal() +
-  theme(axis.text.x = element_text(size = 6,angle = 45, hjust = 1))  # Rotate x-axis labels for readability
+  theme(axis.text.x = element_text(size = 6, angle = 45, hjust = 1))
 
+print(sourcestate)
 ```
+
+```{r}
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/sourcestate.png", plot = sourcestate, width = 8, height = 6, dpi = 300)
+```
+#Based on this graph we see that many of the clinical cases are coming from foods produced in Sweden...interesting. maybe we can use food origin as a predictor as well.
+
 #The left most bar is USA as a whole as some entries do not report the specific state they are from, the rest are states and territories. We see that NY has more cases than other states
@@
 #We see here that the most food isolates for listeria monocytogenes are found in Ready To Eat products, that is a term used by Food Safety and Inspection Service to describe any food products that do not require cooking or sterilization to safely consume.
+
+#plot of food isolates and clinical isolates across 10 years
+```{r}
+library(tidyverse)
+library(lubridate)
+
+# Step 1: Prepare the data — monthly counts
+monthly_counts <- mydata %>%
+  mutate(month = floor_date(`Create date`, "month")) %>%
+  group_by(month) %>%
+  summarize(
+    food_count = sum(food == "Yes"),
+    clinical_count = sum(clinical == "Yes")
+  ) %>%
+  ungroup()
+
+# Step 2: Convert to long format for plotting
+monthly_long <- monthly_counts %>%
+  pivot_longer(cols = c(food_count, clinical_count),
+               names_to = "source_type",
+               values_to = "count")
+
+# Step 3: Plot
+ggplot(monthly_long, aes(x = month, y = count, color = source_type)) +
+  geom_line(size = 1.2) +
+  labs(
+    title = "Monthly Salmonella Isolate Counts by Source",
+    x = "Month",
+    y = "Isolate Count",
+    color = "Source Type"
+  ) +
+  scale_color_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
+  theme_minimal()
+```
+
+```{r}
+Isolatecounts <- ggplot(monthly_long, aes(x = month, y = count, fill = source_type)) +
+  geom_bar(stat = "identity", position = "stack") +
+  labs(
+    title = "Monthly Salmonella Isolate Counts by Source (Stacked Bars)",
+    x = "Month",
+    y = "Isolate Count",
+    fill = "Source Type"
+  ) +
+  scale_fill_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/isolatecounts.png", plot = Isolatecounts, width = 8, height = 6, dpi = 300)
+print(Isolatecounts)
+```

### 2025-04-21 products/manuscript/Manuscript.qmd
updates to manuscript

@@
 ---
-title: "Exploring clinical and non-clinical(food) Listeria monocytogenes isolates "
+title: "Exploring Patterns in Clinical and Non-Clinical Listeria monocytogenes Isolates "
 author: "Alexis Gonzalez"
@@
 
-Will complete at the end
+Listeria monocytogenes is major food born pathogen.
+Though infections are rare, L.monocytogenes has the highest hospitalization rate and death rate of the common food borne illnesses.
+This pathogen can proliferate in a number of hosts and climates.
+Understanding temporal and source relationships may allow the public to be better informed on how to protect themselves.
+This study used whole-genome sequencing data from the NCBI Pathogen Detection database (2010--2024) to explore trends in food, clinical, and environmental isolates of *L.
+monocytogenes*.
+We conducted exploratory data analysis to assess the frequency and geographic distribution of isolates.
+Poisson and logistic regression models were applied to evaluate the effect of seasonality on clinical isolate occurrence.
+33,264 isolates were analyzed, with a majority of them being environmental.
+Geographic analysis concluded that New York state contributed a large percentage of food isolates, and many illness-causing foods originated in Sweden.
+Statistical modeling showed a consistent seasonal trends.
+Random forest models concluded that days of the year were an important factor and Poisson regression allowed for prediction of future clinical isolates based on the number of environmental and food isolates, though the association was weak.
+These findings suggest that utilizing temporal relationships may be effective in predicting potential outbreaks.
 
@@
 Listeria contamination typically happens post-lethality which exposes a ready-to-eat product to a processing environment after a form of lethality treatment has taken place.
-Post lethality processing environments can be deli slicers, submersion in a brine solution, meat encassing, bagging or packaging.
+Post lethality processing environments can be deli slicers, submersion in a brine solution, meat encasing, bagging or packaging.
 The USDA recommends all establishments to take part in additional post lethality treatment such as antimicrobial coatings, pasteurization, or chemical or physical sanitation of equipment.
@@
 \
-This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of a year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products.
+This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products.
 The data source being used to conduct this study is the National Center for Biotechnology Information's Pathogen Detection browser.
@@
 
+Two additional columns were created for exploratory analysis to separate clinical and non clinical isolates from the "Isolation type variable".
+Entries for the "Clinical" column were coded as "Yes" if the entry for "Isolation type" was clinical.
+All non-clinical isolates were coded as "environmental/other" so the "Isolation Source" column was utilized to create the "Food" column.
+All isolates that came from various meat and dairy, and produce food sources were coded as "Yes".
+Additional columns created were "days_since_start" which converted all dates in the selected time period into days since the first date in the data set.
+A column for year, day, month, and season were also created for statistical analysis.
+
+For temporal analysis was conducted using poisson regression.
+All machine learning models were conducted using the tidymodels workflow.
+Specific packages for poisson modeling were parsnip and poissonreg.
+
 {{< pagebreak >}}
@@
 
-## Exploratory/Descriptive analysis
+### Exploratory Data Analysis
+
+Exploratory data analysis started with visualizing the total counts of food, clinical and environmental isolates in the 10 year time period.
+It is seen in *figure* 1 that food isolates account for the smallest proportion of isolated *Listeria monocytogenes* at 10,284.
+Clinical isolates fall slightly ahead at 10,319, with a difference of 35 isolates in the 10 year period.
+Environmental samples dominate the isolate types at 12,777.
+To get a visualization of where food isolates were coming from a bar graph was made showing the location of origin of each food isolate.
+Some isolates only logged "USA" as their origin so those were removed from visualization.
+The majority of food sampled were originate in New York state, followed by Washington, Arizona, Florida, and Texas.
+The 3rd bar graph *figure 3* shows the origin of clinical cases.
+This means the origin of the source of infection, likely a food product.
+An overwhelming percentage of originated in Sweden, followed by Texas and New York.
 
 ```{r}
-knitr::include_graphics(here("results/figures", "foodisolatecount.png"))
+knitr::include_graphics(here("results/figures", "isolatetypes.png"))
 ```
@@
 ```{r}
-
-```
-
-## Basic statistical analysis
-
-```{r}
-
-```
-
-## Full analysis
-
-#Need to insert figures
-
-```{r}
-knitr::include_graphics(here("results/figures", "clinicalxtime.png"))
-
+knitr::include_graphics(here("results/figures","locations.png"))
 ```
@@
 ```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."...................;;;;;;;;;;;;;;;::::::::::::::::P
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
+knitr:: include_graphics(here("results/figures", "sourcestate.png"))
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "isolatecounts.png"))
+```
+
+## Temporal Analysis
+
+*Time of year or "season" as a predictor for clinical Listeria monocytogenes isolates*
+
+The first goal of this analysis is to find any seasonal patterns in *Listeria* infections and *Listeria* contaminated food items.
+
+```{r}
+knitr::include_graphics(here("results/figures", "gam_model.png"))
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "seasonalpreds.png"))
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "poisson_seasonal.png"))
 ```
@@
 {{< pagebreak >}}
+
+## Logistic Regression
 
@@
 
+This study examined temporal and source-based trends in Listeria monocytogenes isolates reported in the United States from 2010 to 2024 using publicly available data from the NCBI Pathogen Detection database.
+Isolates were classified into clinical, environmental, and food-related categories to assess patterns and predictors of clinical infections.
+
+Exploratory analysis showed that environmental samples were the most frequently reported isolate type, which supports previous findings that L.
+monocytogenes is widespread in processing environments and natural reservoirs.
+Clinical and food isolates were nearly equal in count, suggesting a potential link between food exposure and clinical outcomes.
+
+Temporal modeling using Poisson regression revealed a seasonal pattern in clinical isolates, with elevated counts in warmer months---especially summer.
+This pattern may reflect greater bacterial growth in food products stored at warmer temperatures, seasonal shifts in food handling practices, or increased consumption of ready-to-eat items during these months.
+Logistic regression analysis reinforced these findings, highlighting seasonality as a potential predictor of clinical listeriosis.
+
+These results align with prior epidemiological data suggesting that Listeria infections may follow seasonal trends, which has important implications for food safety monitoring and public health messaging.
+Summer could be a critical time for intensified surveillance and intervention efforts.
+
 ## Strengths and Limitations
 
+While the data set contains thousands of isolates, the amount of clinical isolates is quite small due to the rarity of infection.
+This made statistical analysis complicated as some months during the 10 year period experienced 0 cases of listeriosis.
+There also were not many predictors to work with from the raw data.
+Isolation source would have been an association to explore but, the isolation sources are entered manually bu researchers and inspectors so there is no standardization in nomenclature of sources.
+For example, one source can say "hot-dog" while another says "premium hot-dog".
+More spatial analysis could have been done if more entries specify the state of origin.
+Many had just "USA" or NA.
+This is also not standardized for NCBI and can be entered however the submitter pleases.
+
 ## Conclusions
+
+This study identified seasonal patterns in clinical Listeria monocytogenes isolates, with potential implications for timing surveillance and prevention strategies.
+While environmental and food isolates remain a significant reservoir for transmission, more refined metadata and consistent data entry practices are essential for improving future predictive models.
+
+Future research should focus on expanding spatial analysis, refining source categorization through natural language processing, and incorporating additional predictors such as food recall data or consumer behavior trends.
+With enhanced data quality and analytical tools, real-time surveillance systems can be better leveraged to prevent and control outbreaks of listeriosis.
 

### 2025-04-24 code/analysis-code/statistical-analysis.qmd
updated readme files

@@
   "/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", 
-  plot = seasonalpreds, width = 8, height = 6
+  plot = seasonalpreds, width = 8, height = 6)
 
@@
 plot(gam_model)
-ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png", plot = gam_model, width = 8, height = 6, dpi = 300)
 ```

### 2025-04-25 project-review-Nguyen.qmd
Vincent's Review

---
title: Review of Alexis Gonzalez's MADA Project
author: Vincent Nguyen
date: 4/22/2025
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Exploring Patterns in Clinical and Non-Clinical Listeria monocytogenes Isolates

Name of project author(s): Alexis Gonzalez

Name of project reviewer: Vincent Nguyen

## Background, Context and Motivation
### Feedback and Comments

The introductory paragraph greatly contextualizes the project and its aims. I understand the research aims and feel it has a great tangible impact. The introduction is also well researched and gives a scope of how serious listeria infections are. Perhaps a summation of similar previous literature can be implemented to enhance the writing further.

### Summary assessment (PICK ONE, DELETE THE OTHERS)
* strong contextualization and motivation

## Question description
### Feedback and Comments

The research hypothesis is clearly stated and it is clear what variables are important in this analysis. The hypothesis is reasonable and the data source is explained in context with the hypothesis.

### Summary assessment
* question/hypotheses fully clear

## Data description
### Feedback and Comments

The data and its source are well described. The data structure is also explained and some of the filtering done to it. I will note that the README files could use some updating to help explain to people looking at the files.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis
### Feedback and Comments

The data cleaning makes sense and is intuitive. It is not super complex and is explained. Counts and distribution histograms are plotted and help describe the data further. Some of the axes are hard to read, specifically the second figure. Also captions are not attached to the plots. Regardless, eda is intuitve and helps contextualize the analysis.

### Summary assessment
* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

The use of several modeling techniques adds to the robustness of the analysis. Additionally, the models are appropriately used are logically sound. I will note some issues. Lines 39-43 seem to throw up errors. Also, while I am unsure if this is proper workflow, checking some assumptions (like overdispersion for poisson) or adding justifications before model selection could help improve robustness as well.

### Summary assessment
* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

Graphs are aesthetically pleasing but some labels can be hard to read. Additionally, there is a blank picture in "gam_model". Also, while a slight thing, I think choosing a cohesive color palette could make things more consistent throughout the manuscript.

### Summary assessment
* results are very well presented

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

The findings are well discussed and limitations are well acknowledged. I think the results section is missing in my version of the repo but from what I can tell, it seems reasonable. Results from the random forest modeling should be implemented as well.

### Summary assessment
* strong, complete and clear discussion

# Overall project content evaluation
## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

There are some useless files, specificallly, in the assests, products, and results sections of the repository. Additionally, some READMEs are not updated in my revision of the repo.

### Summary assessment
* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

The code has a great presence of comments throughout. I do think some comments could improve in explaining the intent of the code beyond just its function. Additionally, some of the spacing is odd.

### Summary assessment
* decently documented with some gaps

## Reproducibility
### Feedback and Comments

The code is reproducible, however, there is use of absolute paths in some of the figure saving in statistical-analysis.qmd. Additionally, some code straight up does not work but I imagine this will be smoothed over in later revisions. Beyond just a few coding errors, the code looks reasonably reproducible, however, there are redudant files in the code folder. This could make it a bit harder to reproduce more.

### Summary assessment
* small parts not reproducible or required manual intervention

## Thoroughness
### Feedback and Comments

The use of multiple models is a great strength of the analysis. Additionally, the consideration of temporal patterns is also a great way to capture time-based and seasonality trends. I believe the analysis was reasonablly well done and models are logically sound. I do think some assumptions should be checked and possibly residual plots. Also, model comparison metrics would bolster the analysis as well. Also, the discussion could have more fleshed out results regarding the modeling results.

### Summary assessment
* decent level of thoroughness

### 2025-05-02 code/analysis-code/statistical-analysis.qmd
additions

@@
 # Plot
-seasonalpreds <- ggplot(combined_data, aes(x = month, y = clinical_count, color = source)) +
+seasonalpreds <- ggplot(combined_months, aes(x = month, y = clinical_count, color = source)) +
   geom_line() +

### 2025-05-02 code/eda-code/eda.qmd
additions

@@
 
-Load the data.
+#Load the data.
 
@@
 
-Showing a bit of code to produce and save a summary table.
-
-```{r}
-summary_df = skimr::skim(mydata)
-print(summary_df)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-```
-
-# Data exploration through figures
-
 #Making a table for the clinical variable 
@@
 table(mydata$isolation_source)
+```
+
+#Table 1: counts across the 15 year period with range and mean
+
+```{r}
+library(kableExtra)
+
+# Create new year and month columns
+mydata <- mydata %>%
+  mutate(
+    year = year(`Create date`), 
+    month = month(`Create date`, label = TRUE, abbr = TRUE)
+  )
+
+# Count records
+counts <- mydata %>%
+  group_by(year, month, isolation_source) %>%
+  summarize(count = n(), .groups = "drop")
+
+# Summarize by year and source
+yearly_totals <- counts %>%
+  group_by(isolation_source, year) %>%
+  summarize(
+    total = sum(count),
+    highest_month = month[which.max(count)],
+    lowest_month = month[which.min(count)],
+    .groups = "drop"
+  )
+
+# Pivot for final table
+table_data <- yearly_totals %>%
+  pivot_wider(
+    names_from = isolation_source,
+    values_from = c(total, highest_month, lowest_month),
+    names_sep = "_"
+  ) %>%
+  arrange(year)
+
+# Create the styled table
+table_output <- table_data %>%
+  kbl(
+    caption = "Yearly Isolate Summary by Source",
+    col.names = c(
+      "Year", 
+      "Total Clinical", "Total Environmental", "Total Food", 
+      "Highest Month (Clinical)", "Highest Month (Environmental)", "Highest Month (Food)", 
+      "Lowest Month (Clinical)", "Lowest Month (Environmental)", "Lowest Month (Food)"
+    )
+  ) %>%
+  kable_styling(
+    full_width = FALSE, 
+    position = "center", 
+    bootstrap_options = c("striped", "hover", "condensed")
+  )
+```
+
+```{r}
+# Define the file path
+summarytable_file <- here::here("results", "tables", "summarytable.rds")
+
+# Save the styled table as an RDS file
+saveRDS(table_output, file = summarytable_file)
+
 ```
@@
     x = "Location",
-    y = "Count of Yes"
+    y = "Count of Isolates"
   ) + 
@@
   geom_boxplot(fill = "darkblue") +
-  labs(title = "Count of Clinical 'Yes' Across Isolation Sources", 
+  labs(title = "Count of Clinical Isolates Across Isolation Sources", 
        x = "Isolation Source", 
-       y = "Count of Yes") +
+       y = "Count of isolates") +
   theme_minimal() +
@@
        x = "Isolation Source", 
-       y = "Count of Yes") +
+       y = "Count of isolates") +
   theme_minimal() +
@@
   geom_bar(stat = "identity", fill = "darkblue") +
-  labs(title = "Top 10 Food 'Yes' Isolation Sources", 
+  labs(title = "Top 10 Food Isolation Sources", 
        x = "Isolation Source", 
@@
 
-# Step 1: Prepare the data — monthly counts
+#Prepare the data into monthly counts
 monthly_counts <- mydata %>%
@@
     food_count = sum(food == "Yes"),
-    clinical_count = sum(clinical == "Yes")
+    clinical_count = sum(clinical == "Yes"),
+    environmental_count = sum(isolation_source =="environmental")
   ) %>%
@@
 
-# Step 2: Convert to long format for plotting
+#Convert to long format for plotting
 monthly_long <- monthly_counts %>%
-  pivot_longer(cols = c(food_count, clinical_count),
+  pivot_longer(cols = c(food_count, clinical_count, environmental_count),
                names_to = "source_type",
@@
 
-# Step 3: Plot
+#Plot monthly counts
 ggplot(monthly_long, aes(x = month, y = count, color = source_type)) +
@@
   ) +
-  scale_color_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
+  scale_color_manual(values = c("environmental_count" = "forestgreen", "food_count" = "firebrick", "clinical_count"= "darkblue")) +
   theme_minimal()
@@
   ) +
-  scale_fill_manual(values = c("food_count" = "forestgreen", "clinical_count" = "firebrick")) +
+  scale_fill_manual(values = c("environmental_count" = "forestgreen", "food_count" = "firebrick", "clinical_count"= "darkblue")) +
   theme_minimal() +

### 2025-05-02 code/processing-code/processingfile-v1.qmd
additions

@@
 library(here) 
+library(lubridate)
 ```
@@
 # Data loading
-
-Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
-package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
-but it adds typing. You can do it either way.
-
 ```{r}
@@
 summary(rawdata)
-```
-
-Several ways of looking at the data
-
-```{r}
 dplyr::glimpse(rawdata)
@@
 
-```{r}
-unique(rawdata$BioSample)
-```
-```{r}
-unique(rawdata$Location)
-```
-```{r}
-unique(rawdata$`Create date`)
-```
-
 # Cleaning
-
-After inspecting the data,I have dedcided to remove some variables that I will not use.
-This will be serovar, assembly, biosample for now.
-I want to try and make a map so i will keep in isolate names and snp distances and what not.
 
@@
 print(lmono2$`Create date`)
+```
+
+```{r}
+lmono2 <- lmono2 %>%
+  mutate(
+    year = year(`Create date`),
+    month = month(`Create date`, label = TRUE, abbr = TRUE)
+  ) %>%
+  filter(year != 2011)
+
 ```
@@
 
+#Creating an additional variable called "Isolation_source" This is different from the `Isolation Source` variable in that it will include food and not just list it under environmental/other
+```{r}
+lmono_final <- lmono_final %>%
+  mutate(
+    clinical = as.character(clinical),
+    food = as.character(food),
+    isolation_source = case_when(
+      clinical == "Yes" ~ "clinical",
+      food == "Yes" ~ "food",
+      TRUE ~ "environmental"
+    )
+  )
+
+```
+
 ```{r}
@@
 # Save data 
-
-Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
-This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
-However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.
-
-See here for some suggestions on how to store your processed data:
-http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata
-
 ```{r}

### 2025-05-02 products/manuscript/Manuscript.qmd
updated readme files
additions

@@
 library(knitr)
+library(htmltools)
 ```
@@
 
-For temporal analysis was conducted using poisson regression.
-All machine learning models were conducted using the tidymodels workflow.
-Specific packages for poisson modeling were parsnip and poissonreg.
+Temporal analysis was first visualized using
+
+conducted using poisson regression. All machine learning models were conducted using the tidymodels workflow. Specific packages for poisson modeling were parsnip and poissonreg.
 
@@
 An overwhelming percentage of originated in Sweden, followed by Texas and New York.
+
+```{r}
+table_output
+```
 
@@
 The first goal of this analysis is to find any seasonal patterns in *Listeria* infections and *Listeria* contaminated food items.
+In figure 5 we can see that cases of listeriosis typically dip around the 100th day of the year which is early April, and spikes near the 200th to 300th days of the year which are mid July to late October.
+The prediction model forecasts low counts of Listeriosis from the end of 2023 to earlt 2024
 

### 2025-05-03 code/analysis-code/statistical-analysis.qmd
Part 6

--- 2025-04-24 code/analysis-code/statistical-analysis.qmd
@@
 ---
-title: 
-author: 
+title: "Statistical Analysis"
+author: "Alexis Gonzalez"
 date: 
@@
 ---
+
 # loading in packages
+
 ```{r}
@@
 library(poissonreg)
-library(randomForest)
+library(MASS)
 ```
+
 #Load in the data
+
 ```{r}
@@
 
-#Here I will perform some basic statistical analysis.
+# temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
 
-#Lets start with a basic glm of time and food and time and clinical. We will also do some visualizations using a histrogram
-
-#To do this I have to convert my date variable into a numeric variable.
-
-```{r}
-# Load necessary package
-library(lubridate)
-
-"# Sample data
-mydata <- data.frame(
-  date `Create date` = as.Date,
-  outcome = clinical  # Yes = 1, No = 0
-)"
-
-# Recode Yes = 1, No = 0
-mydata$clinical_numeric <- ifelse(mydata$clinical == "Yes", 1, 0)
-
-# Convert date to numeric (days since first observation)
-mydata$days_since_start <- as.numeric(difftime(mydata$`Create date`, min(mydata$`Create date`), units = "days"))
-
-# Logistic regression
-model <- glm(clinical_numeric ~ days_since_start, data = mydata, family = binomial)
-
-# Model summary
-summary(model)
-
-```
-
-# temporal analysis of clinical isolates. Remember that every clinical isolate is like a human case of Listeriosis.
 #For this analysis I will make some changes to my data by separating the time data into seasons.Winter is considered December, January and February; spring is March through May; summer is June through August; and fall or autumn is September through November.
 
+#POISSON MODELING
 ```{r}
-mydata <- mydata %>%
-  drop_na()%>%
-  mutate(
-    `Create date` = as.Date(`Create date`),
-    year = year(`Create date`),
-    day_of_year= yday(`Create date`),
-    week = week(`Create date`),
-    month = month(`Create date`),
-    season = case_when(
-      month %in% c(12, 1, 2)  ~ "Winter",
-      month %in% c(3, 4, 5)   ~ "Spring",
-      month %in% c(6, 7, 8)   ~ "Summer",
-      month %in% c(9, 10, 11) ~ "Fall"
-    ),
-    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall")),
-    `Isolation source` = factor(`Isolation source`)
-  )
-```
-
-#Feature selection
-```{r}
-model1_data <- mydata %>%
-  select(`Isolation type`,year,month,day_of_year,week,season)
-```
-
-#Split the data
-```{r}
-set.seed(1234)
-data_split <- initial_split(model1_data, prop = 0.8, strata = `Isolation type`)
-train_data <-training(data_split)
-test_data <-testing(data_split)
-```
-#Create recipe
-```{r}
-rf_recipe <- recipe(`Isolation type` ~ month + year + day_of_year + week + season, data = train_data) %>%
-  step_dummy(all_nominal_predictors()) %>%
-  step_zv(all_predictors()) %>%
-  step_normalize(all_numeric_predictors())
-```
-
-#Define the model
-```{r}
-rf_mod <-rand_forest(trees = 150, min_n =5) %>%
-  set_mode("classification") %>%
-  set_engine("ranger",importance="impurity")
-```
-Create workflow
-```{r}
-workflow1 <- workflow() %>%
-  add_model(rf_mod)%>%
-  add_recipe(rf_recipe)
-```
-#fit the random forest model
-```{r}
-fit1 <- workflow1 %>%
-  fit(data = train_data)
-```
-#Make predictions for evaluation
-```{r}
-predictions_rf <-fit1 %>%
-  predict(new_data = test_data) %>%
-  bind_cols(test_data)
-
-```
-```{r}
-#set isolation source as a factor
-predictions_rf <- predictions_rf %>%
-  mutate(`Isolation type` = as.factor(`Isolation type`))
-
-```
-
-#Run metrics to evaluate the model
-```{r}
-# Create a metric set for classification
-classification_metrics <- metric_set(accuracy, precision, recall, f_meas)
-```
-
-```{r}
-#  use it on predictions
-metrics_rf <- predictions_rf %>%
-  classification_metrics(truth = `Isolation type`, estimate = .pred_class)
-
-```
-
-```{r}
-metrics_rf <- predictions_rf %>%
-  metrics(truth = `Isolation type`, estimate = .pred_class)
-print(metrics_rf)
-```
-#Plot the results of this model
-
-```{r}
-library(vip)  # for variable importance plot
-
-# Extract the fitted model from the workflow
-rf_fit <- extract_fit_parsnip(fit1)
-
-# Plot variable importance
-vip(rf_fit$fit)
-
-```
-
-#Poisson modeling using month
-
-```{r}
-# Create monthly clinical counts
-monthly_counts <- mydata %>%
+# Create monthly counts for clinical, food, and environmental
+monthly_clinical <- mydata %>% 
   filter(clinical == "Yes") %>% 
@@
   ungroup()
-#filling in gaps with 0
-monthly_counts <- monthly_counts %>%
-  complete(month = seq.Date(min(month), max(month), by = "month"),
-           fill = list(clinical_count = 0))
-#Feature engineering, extracting the month from the data
-monthly_counts <- monthly_counts %>%
-  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+monthly_food <- mydata %>% 
+  filter(food == "Yes") %>% 
+  mutate(month = floor_date(`Create date`, "month")) %>% 
+  group_by(month) %>% 
+  summarize(food_count = n()) %>% 
+  ungroup()
+
+monthly_environmental <- mydata %>% 
+  filter(isolation_source == "environmental") %>% 
+  mutate(month = floor_date(`Create date`, "month")) %>% 
+  group_by(month) %>% 
+  summarize(environmental_count = n()) %>% 
+  ungroup()
+
+# Combine all counts into one dataset
+monthly_counts_combined <- monthly_clinical %>%
+  full_join(monthly_food, by = "month") %>%
+  full_join(monthly_environmental, by = "month")
+
+# View the combined dataset
+head(monthly_counts_combined)
 
@@
 
+#TEMPORAL x CLINCAL
+
 ```{r}
-library(tidymodels)
-library(parsnip)
-#Splitting the data 
+# Filling in gaps with 0 for all counts
+monthly_counts_combined <- monthly_counts_combined %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(clinical_count = 0, food_count = 0, environmental_count = 0))
+
+# Feature engineering: extracting the month from the data
+monthly_counts_combined <- monthly_counts_combined %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+# Splitting the data
 set.seed(123)
-data_split2 <- initial_time_split(monthly_counts, prop = 0.8)
+data_split2 <- initial_time_split(monthly_counts_combined, prop = 0.8)
 train_data2 <- training(data_split2)
@@
 
-```
-```{r}
-#create a recipe 
-poisson_recipe <- recipe(clinical_count ~ month_name, data = train_data2) %>%
+# Create a recipe for clinical_count
+poisson_recipe1 <- recipe(clinical_count ~ month_name, data = train_data2) %>%
   step_dummy(all_nominal_predictors())
 
-```
-```{r}
-#Select the model
-poisson_model <- 
-  poisson_reg(
+# Select Poisson model
+poisson_model <- poisson_reg(
   mode = "regression",
@@
   mixture = NULL,
-    engine ="glm"
+  engine = "glm"
 )
 
-```
-```{r}
-#Create a workflow 
+# Create a workflow
 library(poissonreg)
-poisson_wf <-workflow() %>%
+poisson_wf1 <- workflow() %>%
   add_model(poisson_model) %>%
-  add_recipe(poisson_recipe)
-```
-```{r}
-#fit the model
-poisson_fit <- poisson_wf %>%
-  fit(data = train_data2)
-```
-```{r}
-#make predictions
-predictions <- predict(poisson_fit, new_data = train_data2) %>%
+  add_recipe(poisson_recipe1)
+
+# Fit the model
+poisson_fit1 <- poisson_wf1 %>% fit(data = train_data2)
+
+# Make predictions
+predictions1 <- predict(poisson_fit1, new_data = train_data2) %>%
   bind_cols(train_data2)
-metrics(predictions, truth = clinical_count, estimate =.pred)
-```
-#Visualize results
-```{r}
-poisson_seasonal <- ggplot(predictions, aes(x = month)) +
-  geom_line(aes(y = clinical_count), color = "black", linetype = "dashed") +
-  geom_line(aes(y = .pred), color = "blue") +
-  labs(title = "Predicted vs Actual Monthly Clinical Isolate Counts",
-       y = "Clinical Count")
-print(poisson_seasonal)
-ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/poisson_seasonal.png", plot = poisson_seasonal, width = 8, height = 6, dpi = 300)
-```
-#Predict future isolate counts based on seasonality
-```{r}
-last_month <-max(monthly_counts$month)
-```
 
-```{r}
-future_months <- tibble(
-  month = seq.Date(from = last_month %m+% months(1),by = "month", length.out = 12)
-)%>%
-  mutate(month_name = factor(month(month, label = TRUE), ordered= FALSE))
-```
+# Evaluate performance
+library(yardstick)
+metric_table <- metrics(predictions1, truth = clinical_count, estimate = .pred)
+save_table_location <- here::here("results","tables","metric_table.rds")
+saveRDS(metric_table, file = save_table_location)
+# Predict future clinical isolate counts
+last_month <- max(monthly_counts_combined$month)
 
-```{r}
-#make predictions
-future_predictions <- predict(poisson_fit, new_data = future_months)%>%
-  bind_cols(future_months)
-```
-```{r}
-#visualize
-# Combine historical + future for plotting
-combined_months<- bind_rows(
-  monthly_counts %>% select(month, clinical_count) %>% mutate(source = "Actual"),
-  future_predictions %>% rename(clinical_count = .pred) %>% mutate(source = "Forecast")
-)
-
-# Plot
-seasonalpreds <- ggplot(combined_data, aes(x = month, y = clinical_count, color = source)) +
-  geom_line() +
-  labs(title = "Actual and Forecasted Clinical Isolate Counts",
-       y = "Clinical Count",
-       x = "Month") +
-  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
-  theme_minimal()
-print(seasonalpreds)
-ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", plot = seasonalpreds, width = 8, height = 6, dpi = 300)
-```
-
-```{r}
-
-# 1. Get the last month from your historical data
-last_month <- max(monthly_counts$month)
-
-# 2. Create the next 12 months
-future_months <- tibble(
+future_months1 <- tibble(
   month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
@@
 
-# 3. Make predictions using your Poisson model (or other model)
-future_predictions <- predict(poisson_fit, new_data = future_months) %>%
-  bind_cols(future_months)
+# Future predictions
+future_predictions1 <- predict(poisson_fit1, new_data = future_months1) %>%
+  bind_cols(future_months1)
 
-# 4. Combine historical and forecast data
-combined_months <- bind_rows(
-  monthly_counts %>% 
-    select(month, clinical_count) %>% 
-    mutate(source = "Actual"),
-  future_predictions %>% 
-    rename(clinical_count = .pred) %>% 
-    mutate(source = "Forecast")
+# Combine actual and forecasted for plotting
+combined_months1 <- bind_rows(
+  monthly_counts_combined %>% select(month, clinical_count) %>% mutate(source = "Actual"),
+  future_predictions1 %>% rename(clinical_count = .pred) %>% mutate(source = "Forecast")
 )
 
-# 5. Plot
-seasonalpreds <- ggplot(combined_months, aes(x = month, y = clinical_count, color = source)) +
-  geom_line(size = 1.2) +
-  labs(
-    title = "Actual and Forecasted Clinical Isolate Counts",
-    y = "Clinical Count",
-    x = "Month"
-  ) +
+# Plot actual vs forecast
+seasonalpreds <- ggplot(combined_months1, aes(x = month, y = clinical_count, color = source)) +
+  geom_line() +
+  labs(title = "Actual and Forecasted Clinical Isolate Counts",
+       y = "Clinical Count", x = "Month") +
   scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_minimal()
 
-# 6. Show plot
 print(seasonalpreds)
 
-# 7. Save to file
-ggsave(
-  "/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/seasonalpreds.png", 
-  plot = seasonalpreds, width = 8, height = 6)
+# Extract the fitted glm model from the workflow
+poisson_fit1_glm <- extract_fit_engine(poisson_fit1)
+
+# Calculate deviance residuals
+deviance_residuals_clinical <- residuals(poisson_fit1_glm, type = "deviance")
+
+# Calculate Pearson residuals
+pearson_residuals_clinical <- residuals(poisson_fit1_glm, type = "pearson")
+
+# Plot deviance residuals
+
+residuals_df <- train_data2 %>%
+  mutate(
+    deviance_residual = deviance_residuals_clinical,
+    pearson_residual = pearson_residuals_clinical,
+    fitted = fitted(poisson_fit1_glm)
+  )
+
+# Deviance residual plot
+ggplot(residuals_df, aes(x = fitted, y = deviance_residual)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
+  labs(title = "Deviance Residuals vs Fitted Values",
+       x = "Fitted Values", y = "Deviance Residuals") +
+  theme_minimal()
+
+# Pearson residual plot
+ggplot(residuals_df, aes(x = fitted, y = pearson_residual)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
+  labs(title = "Pearson Residuals vs Fitted Values",
+       x = "Fitted Values", y = "Pearson Residuals") +
+  theme_minimal()
+
+# Estimate dispersion parameter
+# Dispersion = sum of squared Pearson residuals / residual degrees of freedom
+pearson_chisq <- sum(residuals_df$pearson_residual^2)
+resid_df <- poisson_fit1_glm$df.residual
+dispersion_estimate <- pearson_chisq / resid_df
+
+print(paste("Estimated dispersion:", round(dispersion_estimate, 3)))
 
@@
 
-```{r}
-library(mgcv)
-gam_model <- gam(clinical_count ~ s(day_of_year), family = poisson(), data = daily_counts)
-plot(gam_model)
-```
+#TEMPORAL X FOOD
 
 ```{r}
-# Load necessary library
-library(mgcv)
-library(gratia) # for tidy ggplot GAM visualizations
+# Filling in gaps with 0 for food_count
+monthly_counts_combined <- monthly_counts_combined %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(food_count = 0))
 
-# Example: Create ggplot-style plot from your GAM model
-gam_plot <- draw(gam_model)  # `gam_model` is your fitted model
+# Feature engineering: extracting the month from the data
+monthly_counts_combined <- monthly_counts_combined %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
 
-# Now save it
-ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/gam_model.png",
-       plot = gam_plot, width = 8, height = 6, dpi = 300)
+# Splitting the data
+set.seed(123)
+data_split2 <- initial_time_split(monthly_counts_combined, prop = 0.8)
+train_data2 <- training(data_split2)
+test_data2  <- testing(data_split2)
+
+# Create a recipe for food_count
+poisson_recipe_food <- recipe(food_count ~ month_name, data = train_data2) %>%
+  step_dummy(all_nominal_predictors())
+
+# Select Poisson model
+poisson_model_food <- poisson_reg(
+  mode = "regression",
+  penalty = NULL,
+  mixture = NULL,
+  engine = "glm"
+)
+
+# Create a workflow
+poisson_wf_food <- workflow() %>%
+  add_model(poisson_model_food) %>%
+  add_recipe(poisson_recipe_food)
+
+# Fit the model
+poisson_fit_food <- poisson_wf_food %>% fit(data = train_data2)
+
+# Make predictions
+predictions_food <- predict(poisson_fit_food, new_data = train_data2) %>%
+  bind_cols(train_data2)
+
+# Evaluate performance
+library(yardstick)
+metric_table2 <- metrics(predictions_food, truth = food_count, estimate = .pred)
+save_table_location <- here::here("results","tables","metric_table2.rds")
+saveRDS(metric_table2, file = save_table_location)
+
+# Predict future food isolate counts
+last_month <- max(monthly_counts_combined$month)
+
+future_months_food <- tibble(
+  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
+) %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+# Future predictions
+future_predictions_food <- predict(poisson_fit_food, new_data = future_months_food) %>%
+  bind_cols(future_months_food)
+
+# Combine actual and forecasted for plotting
+combined_months_food <- bind_rows(
+  monthly_counts_combined %>% select(month, food_count) %>% mutate(source = "Actual"),
+  future_predictions_food %>% rename(food_count = .pred) %>% mutate(source = "Forecast")
+)
+
+# Plot actual vs forecast
+pmodel_food <- ggplot(combined_months_food, aes(x = month, y = food_count, color = source)) +
+  geom_line() +
+  labs(title = "Actual and Forecasted Food Isolate Counts",
+       y = "Food Count", x = "Month") +
+  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
+  theme_minimal()
+
+# Print and save the plot
+print(pmodel_food)
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/pmodel_food.png",
+       plot = pmodel_food, width = 8, height = 6, dpi = 300)
+
+# Extract the fitted glm model
+poisson_fit_food_glm <- extract_fit_engine(poisson_fit_food)
+
+# Calculate deviance and Pearson residuals
+deviance_residuals_food <- residuals(poisson_fit_food_glm, type = "deviance")
+pearson_residuals_food <- residuals(poisson_fit_food_glm, type = "pearson")
+
+# Create residuals data frame
+residuals_df_food <- train_data2 %>%
+  mutate(
+    deviance_residual = deviance_residuals_food,
+    pearson_residual = pearson_residuals_food,
+    fitted = fitted(poisson_fit_food_glm)
+  )
+
+# Deviance residual plot
+ggplot(residuals_df_food, aes(x = fitted, y = deviance_residual)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
+  labs(title = "Deviance Residuals vs Fitted Values (Food)",
+       x = "Fitted Values", y = "Deviance Residuals") +
+  theme_minimal()
+
+# Pearson residual plot
+ggplot(residuals_df_food, aes(x = fitted, y = pearson_residual)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
+  labs(title = "Pearson Residuals vs Fitted Values (Food)",
+       x = "Fitted Values", y = "Pearson Residuals") +
+  theme_minimal()
+
+# Estimate dispersion
+pearson_chisq_food <- sum(residuals_df_food$pearson_residual^2)
+resid_df_food <- poisson_fit_food_glm$df.residual
+dispersion_estimate_food <- pearson_chisq_food / resid_df_food
+
+print(paste("Estimated dispersion (food model):", round(dispersion_estimate_food, 3)))
 
@@
 
+#TEMPORAL X ENVIRONMENTAL
+
 ```{r}
+# Filling in gaps with 0 for environmental_count
+monthly_counts_combined <- monthly_counts_combined %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(environmental_count = 0))
 
-#  Split the data into training and testing sets
+# Feature engineering: extracting the month from the data
+monthly_counts_combined <- monthly_counts_combined %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+# Splitting the data
 set.seed(123)
-data_split <- initial_split(monthly_counts, prop = 0.8)
-train_data <- training(data_split)
-test_data <- testing(data_split)
+data_split2 <- initial_time_split(monthly_counts_combined, prop = 0.8)
+train_data2 <- training(data_split2)
+test_data2  <- testing(data_split2)
 
-#  Define the recipe (clinical_count predicted by food_count)
-poisson_recipe <- recipe(clinical_count ~ food_count, data = train_data)
+# Create a recipe for environmental_count
+poisson_recipe_env <- recipe(environmental_count ~ month_name, data = train_data2) %>%
+  step_dummy(all_nominal_predictors())
 
-# Define the Poisson regression model
-poisson_model <- poisson_reg(mode = "regression", engine = "glm")
+# Select Poisson model
+poisson_model_env <- poisson_reg(
+  mode = "regression",
+  penalty = NULL,
+  mixture = NULL,
+  engine = "glm"
+)
 
 # Create a workflow
-poisson_wf <- workflow() %>%
-  add_recipe(poisson_recipe) %>%
-  add_model(poisson_model)
+poisson_wf_env <- workflow() %>%
+  add_model(poisson_model_env) %>%
+  add_recipe(poisson_recipe_env)
 
 # Fit the model
-poisson_fit <- poisson_wf %>%
-  fit(data = train_data)
+poisson_fit_env <- poisson_wf_env %>% fit(data = train_data2)
 
-# Make predictions on the training data
-predictions <- predict(poisson_fit, new_data = train_data) %>%
-  bind_cols(train_data)
+# Make predictions
+predictions_env <- predict(poisson_fit_env, new_data = train_data2) %>%
+  bind_cols(train_data2)
 
-#  Evaluate the model 
-isolate_metrics <- metrics(predictions, truth = clinical_count, estimate = .pred)
-print(isolate_metrics)
+# Evaluate performance
+library(yardstick)
+metric_table3 <- metrics(predictions_env, truth = environmental_count, estimate = .pred)
+save_table_location <- here::here("results","tables","metric_table3.rds")
+saveRDS(metric_table3, file = save_table_location)
+# Predict future environmental isolate counts
+last_month <- max(monthly_counts_combined$month)
 
-# Visualize the results
-ggplot(predictions, aes(x = month)) +
-  geom_col(aes(y = clinical_count), fill = "lightgray", alpha = 0.8) +
-  geom_line(aes(y = .pred), color = "blue", size = 1) +
-  labs(
-    title = "Poisson Regression: Predicting Clinical Isolates from Food Isolates",
-    y = "Clinical Isolate Count",
-    x = "Month"
-  ) +
+future_months_env <- tibble(
+  month = seq.Date(from = last_month %m+% months(1), by = "month", length.out = 12)
+) %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+# Future predictions
+future_predictions_env <- predict(poisson_fit_env, new_data = future_months_env) %>%
+  bind_cols(future_months_env)
+
+# Combine actual and forecasted for plotting
+combined_months_env <- bind_rows(
+  monthly_counts_combined %>% select(month, environmental_count) %>% mutate(source = "Actual"),
+  future_predictions_env %>% rename(environmental_count = .pred) %>% mutate(source = "Forecast")
+)
+
+# Plot actual vs forecast
+pmodel_env <- ggplot(combined_months_env, aes(x = month, y = environmental_count, color = source)) +
+  geom_line() +
+  labs(title = "Actual and Forecasted Environmental Isolate Counts",
+       y = "Environmental Count", x = "Month") +
+  scale_color_manual(values = c("Actual" = "black", "Forecast" = "blue")) +
   theme_minimal()
-```
-#Chart showing likelihood
-```{r}
+
+# Print and save the plot
+print(pmodel_env)
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/pmodel_env.png",
+       plot = pmodel_env, width = 8, height = 6, dpi = 300)
+
+#RESIDUALS AND DISPERSION
+
+# Extract the fitted glm model
+poisson_fit_env_glm <- extract_fit_engine(poisson_fit_env)
+
+# Calculate deviance and Pearson residuals
+deviance_residuals_env <- residuals(poisson_fit_env_glm, type = "deviance")
+pearson_residuals_env <- residuals(poisson_fit_env_glm, type = "pearson")
+
+# Create residuals data frame
+residuals_df_env <- train_data2 %>%
+  mutate(
+    deviance_residual = deviance_residuals_env,
+    pearson_residual = pearson_residuals_env,
+    fitted = fitted(poisson_fit_env_glm)
+  )
+
+# Deviance residual plot
+ggplot(residuals_df_env, aes(x = fitted, y = deviance_residual)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
+  labs(title = "Deviance Residuals vs Fitted Values (Environmental)",
+       x = "Fitted Values", y = "Deviance Residuals") +
+  theme_minimal()
+
+# Pearson residual plot
+ggplot(residuals_df_env, aes(x = fitted, y = pearson_residual)) +
+  geom_point() +
+  geom_hline(yintercept = 0, color = "blue", linetype = "dashed") +
+  labs(title = "Pearson Residuals vs Fitted Values (Environmental)",
+       x = "Fitted Values", y = "Pearson Residuals") +
+  theme_minimal()
+
+# Estimate dispersion
+pearson_chisq_env <- sum(residuals_df_env$pearson_residual^2)
+resid_df_env <- poisson_fit_env_glm$df.residual
+dispersion_estimate_env <- pearson_chisq_env / resid_df_env
+
+print(paste("Estimated dispersion (environmental model):", round(dispersion_estimate_env, 3)))
 
@@
 
-#Now I want to do one more advanced statistical analysis.
-This analysis will utilize the counts of clinical and counts of food isolates each year and see if the amount of clinical isolates can be predicted by the amount of food isolates.
-This purpose of this analysis is to see if federal reguulations are doing a good job of preventing illness.
+#Negative Binomial
 
-#put my food and clinical variable into counts and group by year
+#CLINICAL
+```{r}
+# Prepare data
+clinical_counts <- monthly_counts_combined %>%
+  select(month, clinical_count) %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(clinical_count = 0)) %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
+
+# Split
+set.seed(123)
+clinical_split <- initial_time_split(clinical_counts, prop = 0.8)
+clinical_train <- training(clinical_split)
+clinical_test  <- testing(clinical_split)
+
+# Recipe
+clinical_recipe <- recipe(clinical_count ~ month_name, data = clinical_train) %>%
+  step_dummy(all_nominal_predictors())
+
+# Model
+clinical_nb_model <- linear_reg(mode = "regression") %>%
+  set_engine("glm", family = MASS::negative.binomial(theta = 3.5))
+
+# Workflow
+clinical_nb_wf <- workflow() %>%
+  add_model(clinical_nb_model) %>%
+  add_recipe(clinical_recipe)
+
+# Fit
+clinical_nb_fit <- clinical_nb_wf %>%
+  fit(data = clinical_train)
+
+# Training predictions + metrics
+clinical_preds <- predict(clinical_nb_fit, new_data = clinical_train) %>%
+  bind_cols(clinical_train)
+
+metric_table4<-metrics(clinical_preds, truth = clinical_count, estimate = .pred)
+save_table_location <- here::here("results","tables","metric_table4.rds")
+saveRDS(metric_table4, file = save_table_location)
+
+##DISPERSION##
+# Extract the fitted glm object
+nb_fit_engine <- extract_fit_engine(clinical_nb_fit)
+
+# Calculate Pearson residuals
+pearson_resids <- residuals(nb_fit_engine, type = "pearson")
+
+# Residual degrees of freedom
+resid_df <- nb_fit_engine$df.residual
+
+# Dispersion estimate
+dispersion <- sum(pearson_resids^2) / resid_df
+
+# Print
+print(paste("Estimated Dispersion (Negative Binomial):", round(dispersion, 3)))
+
+nmodel_clinical <- ggplot(clinical_preds, aes(x = month)) +
+  geom_line(aes(y = clinical_count, color = "Actual")) +
+  geom_line(aes(y = .pred, color = "Predicted")) +
+  labs(
+    title = "Actual and Predicted Clinical Isolate Counts (Negative Binomial Model)",
+    x = "Month",
+    y = "Clinical Count"
+  ) +
+  scale_color_manual(
+    name = "Source",
+    values = c("Actual" = "black", "Predicted" = "blue")
+  ) +
+  theme_minimal()
+
+# Display the plot
+nmodel_clinical
+
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/nmodel_clinical.png",
+       plot = nmodel_clinical, width = 8, height = 6, dpi = 300)
+```
+
+#FOOD
 
 ```{r}
-monthly_counts2 <- mydata %>%
-  mutate(month = floor_date(`Create date`, "month")) %>%
-  count(month, isolation_source) %>%
-  tidyr::pivot_wider(names_from = isolation_source, values_from = n, values_fill = 0)
-```
+# Prepare data
+food_counts <- monthly_counts_combined %>%
+  select(month, food_count) %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(food_count = 0)) %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
 
-```{r}
-set.seed(123)  # for reproducibility
+# Split
+set.seed(123)
+food_split <- initial_time_split(food_counts, prop = 0.8)
+food_train <- training(food_split)
+food_test  <- testing(food_split)
 
-rf_model <- randomForest(
-  clinical ~ food + environmental,
-  data = monthly_counts2,
-  ntree = 500,          # number of trees
-  importance = TRUE     # track variable importance
-)
-print(rf_model)
+# Recipe
+food_recipe <- recipe(food_count ~ month_name, data = food_train) %>%
+  step_dummy(all_nominal_predictors())
+
+# Model
+food_nb_model <- linear_reg(mode = "regression") %>%
+  set_engine("glm", family = MASS::negative.binomial(theta = 2))
+
+# Workflow
+food_nb_wf <- workflow() %>%
+  add_model(food_nb_model) %>%
+  add_recipe(food_recipe)
+
+# Fit
+food_nb_fit <- food_nb_wf %>%
+  fit(data = food_train)
+
+# Training predictions + metrics
+food_preds <- predict(food_nb_fit, new_data = food_train) %>%
+  bind_cols(food_train)
+
+metrics_table5 <- metrics(food_preds, truth = food_count, estimate = .pred)
+save_table_location <- here::here("results","tables","metrics_table5.rds")
+saveRDS(metrics_table5, file = save_table_location)
+# Extract the fitted glm object from the workflow
+food_nb_engine <- extract_fit_engine(food_nb_fit)
+
+# Calculate Pearson residuals
+pearson_resids_food <- residuals(food_nb_engine, type = "pearson")
+
+# Residual degrees of freedom
+resid_df_food <- food_nb_engine$df.residual
+
+# Dispersion estimate
+dispersion_food <- sum(pearson_resids_food^2) / resid_df_food
+
+# Print the dispersion
+print(paste("Estimated Dispersion (Negative Binomial - Food Model):", round(dispersion_food, 3)))
+
+nmodel_food <- ggplot(food_preds, aes(x = month)) +
+  geom_line(aes(y = food_count, color = "Actual")) +
+  geom_line(aes(y = .pred, color = "Predicted")) +
+  labs(
+    title = "Actual and Predicted Food Isolate Counts (Negative Binomial Model)",
+    x = "Month",
+    y = "Food Count"
+  ) +
+  scale_color_manual(
+    name = "Source",
+    values = c("Actual" = "black", "Predicted" = "blue")
+  ) +
+  theme_minimal()
+
+# Print the plot
+nmodel_food
+
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/nmodel_food.png",
+       plot = nmodel_food, width = 8, height = 6, dpi = 300)
 
 ```
-
-```{r}
-#model_predict_clinical <- lm(clinical ~ food + environmental, data = monthly_counts2)
-#summary(model)
-
+#ENVIRONMENTAL
 ```{r}
 
-```
+# Prepare data
+environmental_counts <- monthly_counts_combined %>%
+  select(month, environmental_count) %>%
+  complete(month = seq.Date(min(month), max(month), by = "month"),
+           fill = list(environmental_count = 0)) %>%
+  mutate(month_name = factor(month(month, label = TRUE), ordered = FALSE))
 
-`
-#Visualzing the future
-```{r}
-# Suppose you created this:
-future_data <- data.frame(
-  month = as.Date(c("2025-02-01","2025-03-01","2025-04-01","2025-05-01", "2025-06-01", "2025-07-01","2025-08-01","2025-09-01","2025-10-01","2025-11-01","2025-12-01")),
-  food = c(300,220, 180, 250,140,122,116,235,96,175,225),
-  environmental = c(350,150, 170, 200,220,140,267,178,155,200,280)
-)
-```
+# Split
+set.seed(123)
+env_split <- initial_time_split(environmental_counts, prop = 0.8)
+env_train <- training(env_split)
+env_test  <- testing(env_split)
 
-```{r}
-# Predict clinical isolates
-future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)
-```
+# Recipe
+env_recipe <- recipe(environmental_count ~ month_name, data = env_train) %>%
+  step_dummy(all_nominal_predictors())
 
-```{r}
-# Predict clinical isolates for future months
-future_data$Predicted_Clinical <- predict(rf_model, newdata = future_data)
+# Model
+env_nb_model <- linear_reg(mode = "regression") %>%
+  set_engine("glm", family = MASS::negative.binomial(theta = 1))
 
-# Combine with past data (no predicted clinicals in the past, so we set it to NA)
-combined_plot_data <- monthly_counts2 %>%
-  select(month, clinical) %>%
-  mutate(Predicted_Clinical = NA,
-         Source = "Historical") %>%
-  bind_rows(
-    future_data %>%
-      transmute(month, clinical = NA, Predicted_Clinical, Source = "Forecast")
-  )
+# Workflow
+env_nb_wf <- workflow() %>%
+  add_model(env_nb_model) %>%
+  add_recipe(env_recipe)
 
-# Filter combined_plot_data for the last 3 years
-recent_data <- combined_plot_data %>%
-  filter(month >= (max(month) %m-% years(3)))
+# Fit
+env_nb_fit <- env_nb_wf %>%
+  fit(data = env_train)
 
-# Plot
-ggplot(recent_data, aes(x = month)) +
-  geom_line(aes(y = Predicted_Clinical, color = Source), size = 1.2, linetype = "dashed") +
-  geom_line(data = recent_data %>% filter(!is.na(clinical)),
-            aes(y = clinical, color = "Observed"), size = 1) +
-  scale_color_manual(values = c("Observed" = "steelblue", "Historical" = "tomato", "Forecast" = "forestgreen")) +
-  labs(title = "Observed and Forecasted Clinical Isolates (Past 3 Years)",
+# Training predictions + metrics
+env_preds <- predict(env_nb_fit, new_data = env_train) %>%
+  bind_cols(env_train)
+
+metrics_table6 <- metrics(env_preds, truth = environmental_count, estimate = .pred)
+save_table_location <- here::here("results","tables","metrics_table6.rds")
+saveRDS(metrics_table6, file = save_table_location)
+# Extract the fitted glm object from the workflow
+env_nb_engine <- extract_fit_engine(env_nb_fit)
+
+# Calculate Pearson residuals
+pearson_resids_env <- residuals(env_nb_engine, type = "pearson")
+
+# Residual degrees of freedom
+resid_df_env <- env_nb_engine$df.residual
+
+# Dispersion estimate
+dispersion_env <- sum(pearson_resids_env^2) / resid_df_env
+
+# Print the dispersion
+print(paste("Estimated Dispersion (Negative Binomial - Environmental Model):", round(dispersion_env, 3)))
+
+nmodel_env <- ggplot(env_preds, aes(x = month)) +
+  geom_line(aes(y = environmental_count, color = "Actual")) +
+  geom_line(aes(y = .pred, color = "Predicted")) +
+  labs(
+    title = "Actual and Predicted Environmental Isolate Counts (Negative Binomial Model)",
     x = "Month",
-       y = "Count",
-       color = "Legend") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+    y = "Environmental Count"
+  ) +
+  scale_color_manual(
+    name = "Source",
+    values = c("Actual" = "black", "Predicted" = "blue")
+  ) +
+  theme_minimal()
+
+# Display the plot
+nmodel_env
+
+ggsave("/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/figures/nmodel_env.png",
+       plot = nmodel_env, width = 8, height = 6, dpi = 300)
 
 ```
-```{r}
-# Create new data (adjust as needed)
-new_data <- data.frame(food = c(200), environmental = c(150))
-
-# Predict clinical isolates
-predict(model_predict_clinical, newdata = new_data)
-
-```
-#Machine Learning since this was just a simple linear regression
-#Random forest
-```{r}
-#set seed and split data
-set.seed(123)
-data_split3 <- initial_split(model_data, prop = 0.8)
-train_data3 <- training(data_split3)
-test_data3 <- testing(data_split3)
-
-```

### 2025-05-03 code/eda-code/eda.qmd
Part 6

@@
 ```
-#Creating an additional variable called "Isolation_source" This is different from the `Isolation Source` variable in that it will include food and not just list it under environmental/other
-```{r}
-mydata <- mydata %>%
-  mutate(
-    clinical = as.character(clinical),
-    food = as.character(food),
-    isolation_source = case_when(
-      clinical == "Yes" ~ "clinical",
-      food == "Yes" ~ "food",
-      TRUE ~ "environmental"
-    )
-  )
-
-```
 
@@
 #Table 1: counts across the 15 year period with range and mean
+```{r}
+# Install once
+webshot::install_phantomjs()
+
+```
 
@@
   )
+print(table_output)
+```
+
+```{r}
+# Save as image
+save_kable(
+  table_output,
+  file = "/Users/alexisgonzalez/Desktop/MADA/Gonzalez-MADA-project/results/tables/yearly_isolate_summary.png"
+)
+
 ```

### 2025-05-03 products/manuscript/Manuscript.qmd
Part 6

--- 2025-04-21 products/manuscript/Manuscript.qmd
@@
 bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
 ---
@@
 library(knitr)
+library(htmltools)
+
 ```
@@
 
--   Alexis Gonzalez $^{1,2,*}$ (ORCID: 0000-0000-1234-5678)
+-   Alexis Gonzalez
 
@@
 Statistical modeling showed a consistent seasonal trends.
-Random forest models concluded that days of the year were an important factor and Poisson regression allowed for prediction of future clinical isolates based on the number of environmental and food isolates, though the association was weak.
-These findings suggest that utilizing temporal relationships may be effective in predicting potential outbreaks.
+Poisson and negative binomial regression models were used to evaluate the predictive value of seasonality and isolate source type on clinical cases.
+While clinical isolates exhibited seasonal trends with higher counts in warmer months, the predictive power of the models was weak (R² ≈ 0.17), and performance was similar across Poisson and negative binomial models.
+These findings suggest that although temporal patterns exist, seasonality alone is insufficient for accurately predicting clinical cases.
+Enhanced data collection and standardization may improve future outbreak prediction efforts.
 
@@
 
-Listeria monocytogenes is a ubiquitous, gram-positive, intracellular pathogen renowned for its adaptability and resilience\[5\].
+Listeria monocytogenes is a ubiquitous, gram-positive, intracellular pathogen renowned for its adaptability and resilience\[3\].
 This bacterium can be found in a wide range of environments such as soil, water, animals, and vegetation.
@@
 The ingestion of Listeria-infected food may lead to listeriosis, a gastrointestinal illness that causes vomiting, diarrhea, and even death.
-It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[10\].
-While immunocompetent individuals may only experience mild gastroenteritis, disseminated infection in the immunocompromised has a lethality rate of \~30%\[8\].  Listeriosis causes 1,600 infections and 260 deaths each year with a 95% hospitalization rate.
+It is of even greater concern in pregnant individuals, as its forceful invasion tactics allow the bacterium to cross the placental barrier, infecting the fetus; this can lead to miscarriages and stillbirths\[6\].
+While immunocompetent individuals may only experience mild gastroenteritis, disseminated infection in the immunocompromised has a lethality rate of \~30%.  Listeriosis causes 1,600 infections and 260 deaths each year with a 95% hospitalization rate\[1\].
 The United States Department of Agriculture lists 6 foods that are at high risks for L.monocytogenes contamination; unpasteurized dairy products, soft cheeses, raw fruits and vegetables (particularly sprouts, enoki mushrooms, and melons), Ready-To-Eat deli meats and hot dogs, refrigerated pate, and refrigerated smoked seafood.
@@
 Post lethality processing environments can be deli slicers, submersion in a brine solution, meat encasing, bagging or packaging.
-The USDA recommends all establishments to take part in additional post lethality treatment such as antimicrobial coatings, pasteurization, or chemical or physical sanitation of equipment.
+The USDA recommends all establishments to take part in additional post lethality treatment such as antimicrobial coatings, pasteurization, or chemical or physical sanitation of equipment\[3\].
 Sanitation tends to be the more cost effective option for many establishments but, due to incorrect sanitation practices and L.monocytogenes robust bio-film formation capabilities sanitation does not always prevent contamination.
 Outbreaks have been seen in a range of products in recent years.
-From 2023 to 2025 there have been over 10 multi-state listeria outbreaks.
+From 2023 to 2025 there have been over 10 multi-state listeria outbreaks\[2\].
 The most recent being an outbreak beginning February 22, 2025 in frozen supplement shakes.
 These shakes were distributed among institutional facilities such as long term care facilities and hospitals.
-This outbreak has resulted in 38 cases, 37 hospitalizations, and 12 deaths so far as this investigation is ongoing as of April 2025.
+This outbreak has resulted in 38 cases, 37 hospitalizations, and 12 deaths so far as this investigation is ongoing as of April 2025\[5\].
 
 \
-This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases.  This study will investigate time of year and number of non-clinical isolates as a predictor for clinical isolates, and food type as a predictor for food isolates , non-clinical isolates that originate in food products.
+This study aims to investigate trends in listeria isolates found in clinical and non-clinical samples in efforts to predict cases. This study will investigate time of year and number of non-clinical isolates as a predictor for clinical isolates , non-clinical isolates being isolates that originate in food products and environmental samples.
 The data source being used to conduct this study is the National Center for Biotechnology Information's Pathogen Detection browser.
@@
 All isolates that came from various meat and dairy, and produce food sources were coded as "Yes".
+A final source column called "isolation_type" was created by assiging "food" to food-\>"Yes" and "clinical" to clinical -\> "Yes", other entries were filled in as "environmental".
 Additional columns created were "days_since_start" which converted all dates in the selected time period into days since the first date in the data set.
@@
 
-For temporal analysis was conducted using poisson regression.
-All machine learning models were conducted using the tidymodels workflow.
-Specific packages for poisson modeling were parsnip and poissonreg.
+Poisson and negative binomial modeling were conducted for temporal analysis using predictors such as season and counts of food/environmental isolates to model monthly clinical isolate counts.
+Overdispersion was assessed using residual deviance and addressed with negative binomial models.
+Model performance was evaluated using RMSE, MAE, and R-squared values.
+All plots were generated using ggplot2
 
@@
 An overwhelming percentage of originated in Sweden, followed by Texas and New York.
+A stacked bar graph was made showing the counts for each isolation source over the 10 year period.
+A large spike in all sources is seen in early 2020 followed by a large dip.
+
+```{r}
+knitr::include_graphics("../../results/tables/yearly_isolate_summary.png")
+
+```
 
@@
 
-## Temporal Analysis
+## Poisson Modeling
 
 *Time of year or "season" as a predictor for clinical Listeria monocytogenes isolates*
-
-The first goal of this analysis is to find any seasonal patterns in *Listeria* infections and *Listeria* contaminated food items.
-
-```{r}
-knitr::include_graphics(here("results/figures", "gam_model.png"))
-```
 
@@
 ```{r}
-knitr::include_graphics(here("results/figures", "poisson_seasonal.png"))
+metric_table <- readRDS(here::here("results", "tables", "metric_table.rds"))
+knitr::kable(metric_table, caption = "Model Performance Metrics for Poisson Regression")
+
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "pmodel2.png"))
+```
+
+```{r}
+metric_table2 <- readRDS(here::here("results", "tables", "metric_table2.rds"))
+knitr::kable(metric_table2, caption = "Model Performance Metrics for Poisson Regression")
+
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "pmodel3.png"))
+```
+
+```{r}
+metric_table3 <- readRDS(here::here("results", "tables", "metric_table3.rds"))
+knitr::kable(metric_table3, caption = "Model Performance Metrics for Poisson Regression")
+
 ```
@@
 
-## Logistic Regression
+## Negative Binomial Model
+
+```{r}
+knitr::include_graphics(here("results/figures", "nmodel_clinical.png"))
+```
+
+```{r}
+metric_table4 <- readRDS(here::here("results", "tables", "metric_table4.rds"))
+knitr::kable(metric_table4, caption = "Model Performance Metrics for Poisson Regression")
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "nmodel_food.png"))
+```
+
+```{r}
+metrics_table5 <- readRDS(here::here("results", "tables", "metrics_table5.rds"))
+knitr::kable(metrics_table5, caption = "Model Performance Metrics for Poisson Regression")
+```
+
+```{r}
+knitr::include_graphics(here("results/figures", "nmodel_env.png"))
+```
+
+```{r}
+metrics_table6 <- readRDS(here::here("results", "tables", "metrics_table6.rds"))
+knitr::kable(metrics_table6, caption = "Model Performance Metrics for Poisson Regression")
+```
 
@@
 
-Temporal modeling using Poisson regression revealed a seasonal pattern in clinical isolates, with elevated counts in warmer months---especially summer.
+Temporal modeling using Poisson regression revealed a seasonal pattern in clinical isolates, with elevated counts in warmer months,especially summer.
 This pattern may reflect greater bacterial growth in food products stored at warmer temperatures, seasonal shifts in food handling practices, or increased consumption of ready-to-eat items during these months.
-Logistic regression analysis reinforced these findings, highlighting seasonality as a potential predictor of clinical listeriosis.
+Though it must be noted that the predictive power was low.
+The R² value was calculated to be 0.17, meaning that there is an effect from season/time of year but that alone is not sufficient in predicting clinical cases.
+Food and environmental models had lower R² values.
+All models all had very high dispersion statistics at 20,34, and 112 for clinical, food, and environmental respectively.
+Negative binomial models produced lower dispersion statistics at 0.923,0.989, 0.94, indicating the model was better fit to the data.
+Unfortunately, predictive performance remained the same, reinforcing the conclusion that seasonality alone is not sufficient for predicting clinical, food, and environmental isolates
 
-These results align with prior epidemiological data suggesting that Listeria infections may follow seasonal trends, which has important implications for food safety monitoring and public health messaging.
+These results loosely align with prior epidemiological data suggesting that Listeria infections may follow seasonal trends, which has important implications for food safety monitoring and public health messaging.
 Summer could be a critical time for intensified surveillance and intervention efforts.
@@
 This is also not standardized for NCBI and can be entered however the submitter pleases.
+It may have also been helpful to expand the timeline from 10 years to 15 or 20 years to allow for more clinical isolate data.
 
@@
 This study identified seasonal patterns in clinical Listeria monocytogenes isolates, with potential implications for timing surveillance and prevention strategies.
+Accurate predictions may allow state and federal governments to take necessary precautions ahead of predicted spikes.
 While environmental and food isolates remain a significant reservoir for transmission, more refined metadata and consistent data entry practices are essential for improving future predictive models.
@@
 
-```{r}
+\[1\] Centers for Disease Control and Prevention. (n.d.). *About listeria infection*. Centers for Disease Control and Prevention. https://www.cdc.gov/listeria/about/index.html
 
-```
+\[2\] Centers for Disease Control and Prevention. (n.d.-b). *Listeria outbreaks*. Centers for Disease Control and Prevention. https://www.cdc.gov/listeria/outbreaks/index.html
+
+\[3\] Farber JM, Peterkin PI.
+Listeria monocytogenes, a food-borne pathogen.
+Microbiol Rev.
+1991 Sep;55(3):476-511.
+doi: 10.1128/mr.55.3.476-511.1991.
+Erratum in: Microbiol Rev 1991 Dec;55(4):752.
+PMID: 1943998; PMCID: PMC372831.
+
+\[4\] *Food Safety and Inspection Service*.
+Listeria Monocytogenes \| Food Safety and Inspection Service.
+(n.d.).
+https://www.fsis.usda.gov/inspection/compliance-guidance/microbial-risk/listeria-monocytogenes
+
+\[5\] Smith, P.
+(2025, February 24). *Nutritional shakes recalled after listeria infections killed 12 and hospitalized dozens dating back to 2018*.
+NBCNews.com.
+https://www.nbcnews.com/news/us-news/nutritional-shakes-recalled-listeria-infections-killed-11-hospitalized-rcna193411
+
+\[6\] Vázquez-Boland JA, Krypotou E, Scortti M.2017.Listeria Placental Infection. mBio8:10.1128/mbio.00949-17.<https://doi.org/10.1128/mbio.00949-17>
+
+\
